---
title: 'Project: Decarbonization Policy for One Country'
author: Jonathan Gilligan
lab_number: 11
lab_date: '2021-04-05'
pubdate: '2020-12-31'
date: '2021-04-05'
bibliography: decarbonization.bib
pdf_url: /files/lab_docs/decarb_project/lab_11/lab_11_instructions.pdf
slug: lab_11_instructions
output:
  blogdown::html_page:
    md_extensions: +tex_math_single_backslash+compact_definition_lists
    toc: yes
---
```{r setup, include=FALSE, cache = FALSE}
knitr::knit_hooks$set(inline = function(x) { knitr:::format_sci(x, 'md')})
knitr::opts_chunk$set(cache = FALSE, echo = TRUE, fig.height=4, fig.width=7)

# This section loads necessary R libraries and sources scripts that define
# useful functions format_md.
#
data_dir = "_data"
script_dir = "_scripts"

if (!dir.exists('data')) dir.create('data')

# Load all of the following packages.
# Install any missing packages.
library(scales)
library(tidyverse)
library(tibble)
library(janitor)
library(knitr)
library(kayadata)

# This sets the default style for ggplot
theme_set(theme_bw(base_size = 15))

# Load basic utility scripts
source('_scripts/format_md.R', chdir = T)
```
# Introduction

This assignment will be more free-form than the lab assignments we have done
earlier in the semester.
For this assignment, you will build on the work you did for the bottom-up and
top-down decarbonization analysis and investigate how decarbonization might
play out in one specific country.
You can work on your own or in a team of two or three for this assignment.
You will choose one country, decide what that country's greenhouse gas
emissions target should be for the year 2050, and then sketch out a plan for
how that country could achieve its goal.

You and your team will turn in a written report (one report
from the whole team is acceptable)
before midnight on Friday April 30.
The report is not meant to be an exhaustive research report. I expect a length
of around 5 pages of doublespaced
text (or equivalently, around 3 pages singlespaced)
per team member, plus appropriate figures, tables, and references.

Unlike previous assignments, you may use RMarkdown, Word, or any other
word-processing mode that you wish when writing your report, but I would like
to you produce PDF output (e.g., if you're writing in Word, save a copy in PDF
format).

# Establishing a Goal

Briefly, look at what your country has pledged to do for emissions reduction
for the 2015 Paris Agreements
(<http://spappssecext.worldbank.org/sites/indc/Pages/INDCHome.aspx> and the
individual Country Briefs at 
<http://spappssecext.worldbank.org/sites/indc/Pages/Content_Brief.aspx>).
The individual country briefs are long and have a lot of material. 
However, there is a single page with the economy-wide mitigation targets,
whcih give a _target year_, a _target type_, and a number for the target type.
The European Union submitted a collective NDC rather than separate commitments
from each member country.

For instance, the United States pledged a target year of 2025, a target type
of "_absolute emission reduction compared to base year (%) 2005_", and
a target amount of 26--28%, which means
that the US has pledged that by 2025, it will have cut its annual greenhouse gas 
emissions 26--28% below the emissions in 2005.
The European Union has pledged a target year of 2030, a target type of
_absolute emissions reduction compared to base year (%) 1990_, and a 
target number of 40%, so it pledges that by 2030 it will have cut its annual
emissions 40% below what it was in 1990.

China chose a different type of target type. It chose a target year of 2030,
a target type of "_reduction in emission intensity per GDP compared to base
year (%) 2005_", and a target number of 60--65%.
This means that China is pledging that by 2030 it will have cut the economic
intensity of greenhouse gas emissions (the Kaya variable _ef_) 60--65% below
what it was in 2005.

A challenge for policy analysts is comparing the different kinds of targets that
different countries specified in their Nationally Destermined Contributions.

To choose your country's goal for 2050 consider what the country is planning to 
do by 2030 or whichever target year it specifies in its 
Nationally Determined Contribution and
then consider how much more the country might reasonably plan to do by
2050 to make a fair contribution to reducing global greenhouse gas emissions
while also taking into account the need for less-developed nations to
grow their economies in order to lift people out of poverty.

For simplicity, I recommend that you focus on carbon dioxide emissions from
burning fossil fuels rather than on all different kinds of greenhouse gases.

There is no right answer, and it would be possible to write a 50-page
paper just on this part, but that is not what I want you to do.
Rather, just present a simple overview of the issues your country faces and how
you would choose a goal.

For instance, you could look back to 1965 (when the data on energy and emissions
in the `kayadata` package begins) and add up the total emissions from your
country and compare it to the total emissions of the world during that period
and then consider how your country's fraction of total emissions compare to your
country's fraction of the total world population or the total world Gross
Domestic Product (GDP).

You may want to refer to the table below, which lists emissions reductions from
2005--2050 for different parts of the world that would meet a global 36%
emissions target in 2050. This table comes from the IPCC's representative
concentration pathway (RCP) database and is based on the pathway called
"RCP 2.6,"" which gives about a two-thirds probability of keeping global warming
below 2&deg; C.
However, this table was produced from an economic model and does not take
account of political and ethical considerations, such as fairness, so you are
not obliged to choose the same goal that this table lists for your country or
its region.

```{r rcp_26, echo=FALSE}
rcp_26 <- tribble(
  ~region, ~target_year, ~ref_year, ~reduction,
  "Australia/New Zealand", 2050, 2005,  0.82,
  "Canada",                2050, 2005,  0.72,
  "China",                 2050, 2005,  0.78,
  "India",                 2050, 2005,  0.73,
  "Japan",                 2050, 2005,  0.66,
  "South Korea",           2050, 2005,  0.67,
  "United States",         2050, 2005,  0.73,
  "Africa",                2050, 2005,  0.28,
  "Latin America",         2050, 2005,  0.40,
  "Middle East",           2050, 2005,  0.32,
  "Southeast Asia",        2050, 2005, -0.17,
  "Western Europe",        2050, 2005,  0.74,
  "World",                 2050, 2005,  0.36
)

rcp_26 %>% mutate(reduction = percent(reduction, 1)) %>% select(-ref_year) %>%
                  kable(digits = 0, align = c("c", "r", "r"),
                        col.names = c("Region", "Year",
                                      str_c("Reduction from ", 
                                            rcp_26$ref_year[1])))
```

# Realizing the Goal

Once you have chosen a goal, you should apply the same kinds of methods that we
used in the bottom-up and top-down decarbonization analyses to estimate what
your conuntry's population and per-capita GDP may be 
**in 2050**, and what
the implications are for energy efficiency and the mixture of fuels that should
supply its energy needs. This is a place to research the country's natural
resources, current energy supply, and opportunities to decarbonize.

For instance, if your country is mountainous with many rivers, you may want to
expand hydroelectricity.
On the other hand, if it is relatively flat, has few rivers, or if all the
major rivers already have dams and generators, then you may want to look at
other sources of energy.
If your country receives a lot of sunshine, you may want to emphasize solar
energy.
Don't forget to think about nuclear energy as well as renewables like
hydroelectricity, wind, and solar.

I am not asking you to give a thorough engineering and economic assessment of
the energy transition, but to give a brief overview of the major opportunities
and obstacles to clean energy in your country and what you think would be the
best strategy for reducing emissions.

# Format

This is a lab report, not a formal research paper, so you do not need to
structure it as a formal paper.
My grading rubric for the written report will be:

* Thoughtful and sensible analysis of goals [30%]
* Thoughtful and sensible analysis of how to realize the goal [30%]
* Good use of the data for your country [20%]
* Organization of the report (do the parts fit together well and tell a clear
  story) [10%]
* Quality of writing [5%]
* Appropriate use of citations and references [5%]

You may choose to turn this in as an RMarkdown document (knitted to PDF, of
course) or a Word document (saved as PDF), or any other format that you have
rendered as PDF.

The key is that you explain clearly how you do your analysis.
You can either use RMarkdown or you can turn in a document written with a
word-processor and refer the reader to R scripts or an RMarkdown document
with "supporting information" that contains the details of your calculations
and quantitative analysis.

You will turn this assignment in by accepting the assignment on
GitHub Classroom, cloning the repository to your own computer, saving your
final products in the repository and committing and pushing them to GitHub.

# Data Sources
The `kayadata` package has data on the Kaya identity variables for 81 countries
and top-down projections for 78 countries.

If you are working on a country that's not one of these, you may want to look
at additional data, such as the International Energy Agency's energy and
emissions analyses.
These reports are extremely expensive to buy, but the Vanderbilt Library 
provides free online access to the World Energy Outlook reports from 1999--2020 
through the Vanderbilt library catalog. The 2020 report is available at this 
link
<https://doi-org.proxy.library.vanderbilt.edu/10.1787/20725302>.
**When you open this page, there will be a button to buy a PDF of the report for
more than $100, but you can ignore that. There are also buttons to download the 
PDF of the full report, or individual chapters, for free. Do not spend your
money buying something the library will give you legally and free.**
You can also get detailed data for energy use in the the OECD nations at
<https://doi-org.proxy.library.vanderbilt.edu/10.1787/379b6cdc-en>.

The U.S. Energy Information Administration
publishes detailed country profiles for a number of countries at 
<https://www.eia.gov/outlooks/ieo/>,
with data on many individual countries at
<https://www.eia.gov/beta/international/>
and brief country and regional analysis reports at
<https://www.eia.gov/beta/international/analysis.cfm>

This list of data sources is not exhaustive and you should feel free to explore
other sources of information, but don't feel that you have to spend hours in the
library or the internet chasing down data. It is fine for this assignment to use
"good enough" data or to substitute estimates if you can't find exact numbers
for what you're looking for.
Just be clear about how you're doing your analysis.
