---
title: Preparation for Lab on Regulating Greenhouse Gas Emissions
author: Jonathan Gilligan
lab_number: 12
lab_date: '2021-04-12'
pubdate: '2020-12-31'
date: '2021-04-12'
pdf_url: /files/lab_docs/regulation_lab/preparation_for_regulation_lab.pdf
slug: lab_12_preparation
output:
  blogdown::html_page:
    md_extensions: +tex_math_single_backslash+compact_definition_lists
    toc: yes
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(lubridate)
# library(xtable)
library(knitr)
library(kableExtra)

game_day = ymd("2021-04-12")
long_game_day = format(game_day, "%A %B %e")
short_game_day = format(game_day, "%b.\ %e")

class_reading_day = ymd("2021-04-05")
long_reading_day = format(class_reading_day, "%A %B %e")
short_reading_day = format(class_reading_day, "%b.\ %e")

```
# Introduction

For the laboratory period on `r long_game_day`, we will do a role-playing 
exercise explore the effectiveness of three different approaches to 
regulating pollution, such as greenhouse gas emissions: command-and-control, 
cap-and-trade, and emissions taxes. 
This reading should serve to prepare you for the exercise.

# Zero-Emissions Baseline

To simplify things for doing this exercise, we will calculate costs and 
benefits with respect to a zero-emissions baseline, where we specify both the
marginal and cumulative (or gross) costs and benefits with respect to a 
condition where there are no emissions. 

In this case, from the perspective of the businesses that you will be playing,
the costs will be the cost of buying permits to emit CO~2~ or the costs of
paying taxes on emissions. 
The benefits to the businesses will be the profits they earn. 
The costs to society will be the social cost of carbon (the damage caused by 
global warming) and the benefit will be the combination of revenues the 
companies generate (this benefits the companies' owners as profit, 
but the revenues also provide wages for workers and spending by the company, 
which stimulates the economy; this total economic activity is what we measure
as the companies' contribution to the nation's gross domestic product)
as well as money received from the players as taxes or payment for emissions 
permits, which is then refunded to the people through some mechanism, such as
a dividend from the government or a reduction in payroll taxes.

# The Exercise

## The Players

To keep groups small, I will divide the class in half and each half will perform 
the same role-playing exercise. For the exercise, I will divide the students 
into three groups:

1. One group will play The Environmental Protections Agency (EPA) and will 
   decide how much to reduce pollution, how many permits to issue, or how 
   much to charge as a pollution tax. 
   
   The EPA's motivation is to produce the best net benefit for society by 
   balancing the costs of reducing greenhouse gas emissions against the 
   benefits of limiting global warming.
2. A second group will play Alpha Electricity: a large power company with a 
   varied portfolio of generating plants including coal, natural gas, and 
   nuclear. 
   
   Alpha's motives are purely to produce the greatest profit for its 
   shareholders regardless of the cost to society of greenhouse gas emissions.
3. A third group will play Beta Industries: a large heavy-industrial 
   conglomerate with many large factories producing steel, aluminum, and 
   petrochemicals such as plastics, paints, and pharmaceuticals. 
   
   Beta's motives are purely to produce the greatest profit for its 
   shareholders regardless of the cost to society of greenhouse gas 
   emissions.

For the purposes of this exercise, we will assume that the EPA can accurately 
estimate the damage that would be caused by global warming, and thus, that it 
can also accurately estimate the social benefit of reducing greenhouse gas 
emissions. However, the EPA cannot accurately assess the costs individual 
companies will incur when they reduce their emissions. This means that the 
EPA's estimates of net benefits (benefits minus costs) is limited by its 
uncertainty about the cost of reducing emissions.

Only the EPA will know the social cost of greenhouse gas emissions, only Alpha 
will know Alpha's cost for reducing emissions, and only Beta will know Beta's 
cost for reducing emissions.

Without regulation, Alpha and Beta would each emit 150 million tons of CO~2~ 
per year. The goal of the exercise is for the EPA to reduce pollution to achieve 
the socially optimal balance between the benefits of economic activity 
(jobs, wealth, etc.) and the harms of pollution. The goal of each firm is to 
maximize their profit regardless of the costs or benefits to society or to its 
competitor.

To keep things simple, emissions cuts will be figured in blocks of 
1 million tons, so a firm can cut emissions by zero, 1 million tons, 
2 million tons, ..., up to a maximum of 15 million tons (cutting emissions 
by 15 million tons means the firm reduces its emissions to zero).

\clearpage
## The Game

The game will have six stages:

1. First, the EPA will gather information on the cost of abating pollution. 
   A representative of the EPA can ask four questions about each company's 
   marginal costs and benefits
   (**I recommend asking questions about the marginal profit for the 
   $n^{\text{th}}$ million tons of pollution:
   "what is your marginal profit for the 6th million tons you emit?"**). 
   A representative of each company will answer the question. 
   The representatives may answer strategically, meaning they may exaggerate the 
   costs. The EPA may take this into account in deciding how to use the answers 
   to estimate the true cost of reducing emissions.
2. Second, the EPA will determine three possible courses of action:
   a) A command and control regulation, which mandates a specific emissions 
      by each firm. Because the Constitution guarantees equality before the law, 
      this regulation must impose the same emissions limit on each 
      firm.^[The real world is a bit more complex, because the government 
      would be allowed to regulate different industries differently, but would 
      face real obstacles if it tried to impose different emissions limits on 
      different companies within the same industry. Because this exercise is 
      limited to just two companies, we simplify by forcing the EPA to impose 
      identical emissions limits on all firms.]
      Based on what it knows about the costs (to the firms) and benefits 
      (to society) of emissions, the EPA will determine how many 
      million tons of emissions to allow (to keep things simple, the EPA 
      should make the amount an even number: 2 million tons, 4 million tons, 
      6 million tons, etc.) and then divide those emissions allowances equally 
      between the two firms.
   b) A cap-and-trade. Under this program, the total emissions cuts would be 
      the same and the EPA will issue permits to emit CO~2~. Each permit will 
      allow the owner to emit 1 million tons of CO~2~. Total CO~2~ emissions are 
      30 million tons minus the emissions cuts the EPA wants to impose. 
      Again, to keep things simple, issue an even number of permits.
      The EPA will give each company an equal number of permits (if the EPA 
      issues 10 permits, it gives 5 to each company).
	c) An emissions-tax program. Under this program, the EPA determines the tax 
		a firm must pay for every million tons of CO~2~ it emits. The firms will 
		then decide on their own how much CO~2~ they want to emit.
3. Third, the firms determine how much CO~2~ they will emit under the 
   command-and-control program (this is easy, because it's the amount the EPA
   allows). 
4. Fourth, permits will be distributed. The EPA will distribute equal numbers 
   of permits to each firm and allow the firms to trade permits with each other. 
5. Fifth, we will impose an carbon tax, at the price set by the EPA in stage 2. 
   The two firms will be free to cut emissions by as much or as little as they 
   want (between 0 and 15 million tons each). 
6. Sixth and finally, we will reveal the details from each step, including the 
   private information each firm has about its costs, and calculate the 
   deadweight loss under each of the three types of regulation. 
   This will let us discuss and analyze the strengths and limitations of each 
   regulatory program.

## Acknowledgements

This exercise was adapted from The Pollution Game, an interactive exercise 
developed by Jay R. Corrigan, Associate Professor of Economics, Kenyon College.  
See J.R. Corrigan, ``The Pollution Game: A Classroom Exercise Demonstrating the 
Relative Effectiveness of Emissions Taxes and Tradable Permits,'' 
_The Journal of Economic Education_ **42**, 70--78 (2011) 
[doi: 10.1080/00220485.2011.536491](https://doi.org/10.1080/00220485.2011.536491)

\clearpage
## Optional Preparation Exercises: 

To prepare for the lab, please do the following exercises. 
**You do not need to turn it in**, but if you are unable to attend the lab 
session on `r long_game_day`, then you should turn it in on 
Brightspace to get credit for this lab):

1. Table 1 lists the marginal profit a company earns by emitting CO~2~. 
   Fill in the blanks in the table to show the total profits it earns at each 
   amount of CO~2~ emissions.

2. What level of CO~2~ emissions would produce the maximum total profit? 
   How much profit would this be?
   \vspace{2cm}

3. If the company is emitting the amount of CO~2~ that would maximize its 
   profits, and then the Environmental Protection Agency requires the company 
   to reduce its emissions by 5 million tons, what is the total cost for the 
   company to comply with the regulation?
   \vspace{2cm}

4. What is the marginal cost to comply with the regulation? 
   (Be careful and consider, if the company cuts 1 million tons, then a second 
   million tons, and so forth, what did it cost the company to make the 
   fifth million-ton cut?)
   \vspace{2cm}

5. If the EPA imposed a tax of \$30 per ton on CO~2~ emissions, complete the 
   table to indicate the new marginal profit and total profit at each level of 
   emission.
   \vspace{2cm}

6. Under a \$30 per ton tax, what level of CO~2~ emissions would produce the 
   maximum total profit? What would its total profit be? How much less is this 
   than the total profit you reported in question 2?
   \vspace{2cm}

7. Table 2 shows the economic value of the marginal 
   environmental harm caused by each additional million tons of CO~2~ emissions. 
   Fill in the blanks to indicate the total environmental harm.

8. The marginal net economic impact is the marginal profit generated by emitting 
   each million tons of CO~2~ minus the marginal environmental harm. This number 
   is the net benefit to society from emitting an additional million tons of 
   CO~2~. If the number is positive, society benefits. If it is negative, society 
   suffers.

    Fill in the marginal and total economic impacts, using the information on 
    environmental harm from Table 2 and the information 
    on marginal profits from Table 1.

9. What is the optimum amount of CO~2~ to emit if we consider the net benefit to 
   society (i.e., the net economic impact).
   \vspace{2cm}

10. If you were going to set a cap on emissions, how many tons would you set the 
    cap at?
   \vspace{2cm}

11. If you were going to set a tax on emissions, how many dollars per ton would 
    you set the tax at?
   \vspace{2cm}

### Tables: Emissions, Profits, and Harms

```{r profit_table}
tax = 30
mp_intercept = 91
mp_slope = 6
mc_intercept = 0
mc_slope = 5

mp_fn = function(emis) {
   mp_intercept - mp_slope * emis
}
mc_fn = function(emis) {
   mc_intercept + mc_slope * emis
}

e_star = round((mp_intercept - mc_intercept) / (mp_slope + mc_slope))
optiumal_tax = mp_fn(e_star)

profits = tibble(
  emissions = seq(1,20), 
  marginal_profit = mp_fn(emissions),
  total_profit = str_c(marginal_profit, 
                       ifelse(is.na(lag(marginal_profit)), "", 
                              str_c(" + ", lag(cumsum(marginal_profit)), 
                                    " = ", cumsum(marginal_profit))
                              )
                       ),
  marginal_taxed = str_c(marginal_profit, " - ", tax, " = ",
                         marginal_profit - tax),
  total_taxed = str_c(marginal_profit - 30,
                       ifelse(is.na(lag(marginal_profit)), "", 
                              str_c(" + ", lag(cumsum(marginal_profit - 30)), 
                                    " = ", cumsum(marginal_profit - 30))
                              )
                      ),
) %>% mutate(across(total_profit:total_taxed,
                ~ifelse(emissions <= 3, ., "")))

column_names <- c("CO~2~ Emissions\n(million tons)",
                    "Marginal profit\n(million dollars)", 
                    "Total profit\n(million dollars)",
                    "Marginal profit\n(million dollars)", 
                    "Total profit\n(million dollars)")

output_format <- ifelse(is_html_output(), "html", "latex")
# message("Output format = ", output_format)

if (output_format == "latex") {
   escape = FALSE
   column_names <- linebreak(column_names) %>%
      str_replace_all("CO~2~", "$\\\\text{CO}_2$")
   profits <- profits %>% 
      mutate(across(everything(), ~str_replace_all(., fixed("$"), "\\$")))
} else {
   lb <- function(x) x
   escape = TRUE
}

kable(profits, format = output_format, align = "crrrr", digits = 0, 
      col.names = column_names,
      booktabs = T, escape = FALSE,
      caption = "**Table 1:** Profits versus emissions: The table lists the marginal profit, in millions of dollars, for each million tons of CO~2~ it emits. For instance, if it emits one million tons, it earns a marginal profit of $85 million. The total profit for _x_ million tons is the sum of the marginal profits for each million tons from 1 to _x_.\n\nFill in the blanks with the total profit the company earns by for each amount of CO~2~ emission.") %>% 
  add_header_above(c(" " = 1, "No Emissions Tax" = 2, 
                     "$30/ton Emissions Tax" = 2)) %>%
  kable_styling(c("bordered","hover"), c("striped", "scale_down"))
```


```{r cost_table}
costs = tibble(
  emissions = seq(1,20), 
  harm = mc_fn(emissions),
  cum_harm = cumsum(harm),
  total_harm = str_c(ifelse(is.na(lag(harm)), "", 
                              str_c(harm, " + ", lag(cum_harm), 
                                    " = ")), 
                              cum_harm),
  profit = mp_fn(emissions),
  cum_profit = cumsum(profit),
  marginal_profit = str_c(profit),
  total_profit = str_c(cum_profit),
  impact = profit - harm,
  cum_impact = cum_profit - cum_harm,
  marginal_impact = str_c(profit, " - ", harm, 
                          " = ", impact),
  total_impact = str_c(cum_profit, " - ", cum_harm, " = ", cum_impact)
) %>% mutate(across(total_harm:total_impact,
                ~ifelse(emissions <= 3, ., ""))) %>%
  select(-profit, -impact, -starts_with("cum_"))

column_names <- c("CO~2~ Emissions\n(million tons)",
                    "Marginal\n(million dollars)", 
                    "Total\n(million dollars)",
                    "Marginal\n(million dollars)",
                    "Total\n(million dollars)",
                    "Marginal\n(million dollars)", 
                    "Total\n(million dollars)")

output_format <- ifelse(is_html_output(), "html", "latex")
# message("Output format = ", output_format)

if (output_format == "latex") {
   column_names <- linebreak(column_names) %>%
      str_replace_all("CO~2~", "$\\\\text{CO}_2$")
   escape = FALSE
   costs <- costs %>% 
      mutate(across(everything(), ~str_replace_all(., fixed("$"), "\\$")))
} else {
   escape = TRUE
}


kable(costs, format = output_format, align = "crrrrrr", digits = 0,
      col.names = column_names, booktabs = T, escape = escape,
      caption = "**Table 2:** Economic impacts of CO~2~ emissions. The table lists the marginal environmental harm, in millions of dollars, for each million tons of CO~2~ released into the environment. For instance, the first million tons of emissions cause $5 million in damage.\n\nThe right pair of columns shows the net economic impact of the profits generated from the emissions (see Table 1) minus the environmental harm.\n\nFill in the blanks with the total environmental harm from CO~2~ emission and\n\ncompute the net economic impact.") %>% 
  add_header_above(c(" " = 1, "Environmental Harm" = 2, 
                     "Profits" = 2,
                     "Net Economic Impact" = 2)) %>%
  kable_styling(c("bordered","hover"), c("striped", "scale_down"))
```
