---
title: Instructions for Bottom-Up Decarbonization Policy Analysis
author: Jonathan Gilligan
lab_number: 10
lab_date: '2021-03-29'
pubdate: '2020-12-31'
date: '2021-03-29'
bibliography: decarbonization.bib
pdf_url: /files/lab_docs/lab_10/lab_10_instructions.pdf
slug: lab_10_instructions
output:
  blogdown::html_page:
    md_extensions: +tex_math_single_backslash+compact_definition_lists
    toc: yes
---
```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE)

is_output_html <- function() {
  is_html <- TRUE
  header <- rmarkdown:::parse_yaml_front_matter(
    readLines(knitr::current_input())
  )
  if ("output.blogdown::html_page" %in% names(header)) {
    is_html <-  TRUE
  } else if ("output" %in% names(header)) {
    if (is.list(header$output)) {
      output <- names(header$output)[1]
    } else {
      output <- header$output[1]
    }
    if (str_detect(output, regex("(pdf|word)_document", ignore_case = TRUE))) {
      is_html <- FALSE
    }
  }
  is_html
}


getOutputFormat <- function() {
  output <- rmarkdown:::parse_yaml_front_matter(
    readLines(knitr::current_input())
    )$output
  if (is.list(output)){
    return(names(output)[1])
  } else {
    return(output[1])
  }
}

data_dir = "_data"
script_dir = "_scripts"

library(magrittr)
library(tidyverse)
library(scales)
library(lubridate)
library(knitr)
library(broom)
library(kayadata)

theme_set(theme_bw(base_size = 15))

if (!dir.exists(data_dir)) dir.create(data_dir)

source(file.path(script_dir, "utils.R"), chdir = T)

due_date_1 <- lubridate::date("2021-04-14")

long_due_date_1 <- stamp("Monday, March 1")(due_date_1)
short_due_date_1 <- stamp("Mar. 1")(due_date_1)

current_year <- max(get_kaya_data("United States")$year, na.rm = T)
after_current_year <- current_year + 1

ref_year <- 2005
target_year  <- 2050
target_reduction <- 0.80

eia_report_year <- max(get_top_down_values("United States")$year, na.rm = T)
donut_year <- max(kayadata::get_fuel_mix("United States")$year, na.rm = T)

IEO_url <- "http://www.eia.gov/forecasts/ieo/"
EIA_table_url <- "http://www.eia.gov/oiaf/aeo/tablebrowser"

# bottom_up_country_assignments <- bind_rows(
#   tibble(class = "Undergraduates", 
#          region = c("World", "United States", "China",
#                     "South Korea", "Canada", "India",
#                     "Sweden")),
#   tibble(class = "Grad Students", 
#          region = c("World", "United States", "China",
#                      "India", "Brazil"))
# )

rcp_26 <- tribble(
  ~region, ~target_year, ~ref_year, ~reduction,
  "Australia/New Zealand", 2050, 2005,  0.82,
  "Canada",                2050, 2005,  0.72,
  "China",                 2050, 2005,  0.78,
  "India",                 2050, 2005,  0.73,
  "Japan",                 2050, 2005,  0.66,
  "South Korea",           2050, 2005,  0.67,
  "United States",         2050, 2005,  0.73,
  "Africa",                2050, 2005,  0.28,
  "Latin America",         2050, 2005,  0.40,
  "Middle East",           2050, 2005,  0.32,
  "Southeast Asia",        2050, 2005, -0.17,
  "Western Europe",        2050, 2005,  0.74,
  "World",                 2050, 2005,  0.36
)

```
# Introduction

The purpose of this lab is to get a sense of the challenges to cutting 
emissions significantly by analyzing a two representative emissions-reduction
policies.
These policy analyses will follow the methods Roger Pielke used in 
Chapters 3--4 of _The Climate Fix_.

I encourage you to work with a partner on this lab, but you should write up
your own lab report individually.

## Data Resources

To make things simple for you, I have prepared an interactive web application, 
available at <https://ees3310.jgilligan.org/decarbonization/}, with almost all 
the data you will need for this project. 
It contains historical data on population, GDP, energy consumption, and 
CO~2~ emissions for many countries and regions of the world. 

I have also provided an R package that you can install on your own computer 
through R Studio:
```r
library(pacman)
p_load_("kayadata")
```

Finally, there is an experimental version of the interactive web application
that you can install and run on your computer using RStudio, but it is still
experimental and may not work perfectly.
You can install it in RStudio like this:
```r
library(pacman)
p_install_gh("jonathan-g/kayatool")
```
and then you can run the application like this:
```r
library(kayatool)
launch_kayatool()
```
**Note:** you should not put `launch_kaya_tool()` in RMarkdown documents,
like your lab report, because it's for launching an interactive web application,
which is not something that makes sense in a document. Also, launching an 
interactive application when you knit your report will prevent the report from 
knitting correctly.

### Using the interactive web application:

To use the decarbonization web app, start by selecting a country, region, or 
group of countries on the 
left-hand control panel. Then you can set the parameters for your policy goals: 
The target year for accomplishing the emissions reductions, the reductions you 
hope to achieve for the country or region, and the reference year. 

For instance, if your 
goal is for emissions in `r target_year` to be 
`r label_percent(accuracy=1)(target_reduction)` less than they were in 
`r ref_year`, you would 
put 
`r target_year` for the target year, 
`r label_percent(accuracy=1)(target_reduction)` for the emissions reduction, 
and `r ref_year` for the reference year.  
If you want to indicate a growth in emissions, rather than a reduction, 
just enter a negative number for the emissions reduction.

You can also select what year to use for starting the calculation of bottom-up 
trends in the Kaya-identity parameters 
population _P_, 
per-capita gross-domestic product _g_, 
energy intensity of the economy _e_, 
and carbon intensity of the energy supply _f_. 
When you calculate decarbonization rates in this homework project, you will be 
focusing on the carbon intensity of the economy, which is given by the 
product _ef_.

After you have set the parameters you want, the bottom of the left panel will 
show a "Bottom-up Analysis" table that shows the average percentage growth 
rates for the Kaya parameters, their actual values in 2017, and the bottom-up 
projections for what their values will be in the target year (2050 by default).

The tabs on the right-hand side of the web page show:

* "**Trends:**" shows historical trends and the calculated growth rate for the 
  Kaya parameters. You select a variable (_P_, _g_, _e_, _f_, or various 
  multiples _ef_, $G = Pg$, $E = Pge$, or $F = Pgef$) The app shows two graphs: 
  on the right, the value of the parameter and on the left, 
  the natural logarithm of the parameter, which we use to calculate 
  percentage growth rates. 
  The graphs show the points that are used in calculating the trends in 
  darker red and the points not used in the trend calculation in lighter red. 
  If you change the starting year on the left-hand panel, you will see the 
  colors of the dots change to reflect this.

  The trend is shown in black on the left-hand graph. If the quantity is 
  changing at a steady rate, the data points will follow a straight line 
  (the trend line). Sometimes you will see that the variables _e_ and _f_ do 
  not seem to be changing at a steady rate, but the product _ef_ is. 
  Explore the trends for the different variables and notice which seem to be 
  following a steady growth or reduction and which do not.

  If you hold the mouse pointer over a data point on either graph, a tool-tip 
  will pop up showing the value of that variable in that year.

* "**Calculations**" shows the steps for you to follow in this 
  lab exercise.

* "**Implied Decarbonization**" shows the historical trend in the carbon 
  intensity of the economy (_ef_) and the implied future changes in order to 
  meet the policy goal that you set.

* "**Energy Mix**" shows the mixture of energy sources (coal, natural gas, oil, 
  nuclear, and renewables) that provide the country or region's energy supply. 
  From this page, you can download the energy mix for the country you're 
  looking at as a text file, using comma-separated value (csv) format, which 
  you can read into R, Excel, or any other common data anlysis program.

* "**Historical**" shows a table of historical values for the different Kaya 
  parameters. This is a convenient place to look up the exact numbers for your 
  country in a particular year.
  This sheet also has a download button that lets you download the data in a 
  `.csv` file.

# Background and Context

The basic framework for your analysis will be the Kaya identity:
$$
F = P \times g \times e \times f,
$$
where 
_F_ is the CO~2~ emissions (in million metric tons of carbon per year), 
_P_ is the population (in billion people), 
_g_ is the per-capita GDP (in thousands of dollars per person per year), 
_e_ is the energy intensity of the economy (in quads per trillion dollars of GDP), 
and _f_ is the carbon intensity of the energy supply (in million metric tons of 
carbon dioxide per quad).^[One metric ton = 1000 kg = 1.1 English tons = 2200 pounds] 
A quad means one quadrillion British thermal units (BTU) of energy. 
One quad is approximately equal to 8 billion gallons of gasoline or 
36 million tons of coal. 
It is roughly equal to the electricity used by 
26 million homes in a year, 
or the amount of electricity generated by 
15 nuclear power plants in a year.

We will also focus on the carbon intensity of the economy (in metric tons of 
CO~2~ emissions per million dollars of GDP), which 
equals $e \times f$.^[Note that _e_ is in units of quads per trillion dollars 
of GDP and _f_ is in units of million metric tons of CO~2~ per quad, 
so if you multiply the units you get million metric tons of CO~2~ per 
trillion dollars of GDP, which equals metric tons of CO~2~ per 
million dollars of GDP.]

## Growth Rates and Trends

We will assume that all of the rates of change in the growth and 
decarbonization trends we are studying will be constant from year to year.
A constant percentage rate of change implies that the quantity follows
an exponential growth function, 
so if you know the values for _P_, _g_, _e_, and _f_ in `r current_year`, 
then at some future year _y_:

```{r lefteqn_for_mathjax, echo=FALSE, eval=TRUE, results="asis", message=FALSE, warning=FALSE}
# Insert missing definitions of \lefteqn and \intertext for MathJax if the 
# output is a variant on HTML.
message("Output format = ", getOutputFormat())
if(is_output_html()) {
  message("Defining lefteqn")
  cat("$\\def\\lefteqn#1{\\rlap{\\displaystyle{#1}}}$\n")
  cat("$\\def\\alintertext#1{\\rlap{\\displaystyle{\\text{#1}}}}$\n")
} else {
  message("Not defining lefteqn")
  cat("")
}
```
$$
\begin{aligned}
  P(y) &= P(`r current_year`) \times \exp(r_P (y - `r current_year`)),\\
  g(y) &= g(`r current_year`) \times \exp(r_g (y - `r current_year`)),\\
  e(y) &= e(`r current_year`) \times \exp(r_e (y - `r current_year`)),\\
  \alintertext{and}
  f(y) &= f(`r current_year`) \times \exp(r_f (y - `r current_year`)),\\
\end{aligned}
$$
where _r~P~_ is the growth rate of the population, _r~g~_ is the growth rate of the per-capita GDP, etc.
Increasing energy efficiency and/or decarbonization of the energy supply mean that _r~e~_ and/or _r~f~_ are negative.

------

**Remember that you have to divide percentages by 100 to get the rates for 
these equations: if _r_ is 3%, you use 0.03, not 3.0 in the equations.**

**In your math classes and on your calculator, you have probably seen the 
exponential function exp(_x_) written as _e^x^_, where _e_ is the base of the 
natural logarithm (_2.718..._). 
But since I am using the letter _e_ to represent the energy intensity of the 
economy 
(the energy consumption divided by the GDP), 
I am writing it as exp(_x_) so you won't get confused by two different 
meanings of "_e_." Also, in R the exponential function is `exp()`.**

------

Because of the properties of the exponential function, when you multiply two or 
more quantities together, the rate of change of the product is the sum of the 
rates of change of each of the quantities:
$$
\begin{aligned}
  \mathrm{GDP}(y) &= P(y)\times g(y)\\
    &= P(`r current_year`) \times \exp(r_p (y - `r current_year`)) 
    \times g(`r current_year`) \times \exp(r_g (y - `r current_year`))\\
    &= P(`r current_year`)\times g(`r current_year`)
    \times \exp((r_P + r_g) (y - `r current_year`))\\
    \alintertext{so}
  r_{\mathrm{GDP}} &= r_{P\times g} = r_P + r_g.
\end{aligned}
$$

The web app does these calculations so you can check your results.
So that errors in the first parts of a problem don't cascade through the whole 
exercise, you should work the problems exercises with RMarkdown and compare 
your work to the "Bottom-up Analysis" table to make sure you know how to do it.


# The Assignment

For this lab, you will do a bottom-up analysis for a policy to reduce the 
greenhouse gas emissions for the OECD (the Organization
for Economic Cooperation and Development, a group of highly industrialized 
countries) `r label_percent(accuracy=1)(target_reduction)` below 
`r ref_year` levels by the year `r target_year`

```{r country_table, include=FALSE, eval = FALSE}
bottom_up_country_assignments %>% group_by(class) %>%
  summarize(regions = str_c(region, collapse = ", ")) %>%
  kable(align = "ll")
```

For the bottom-up analysis, use the Kaya Identity to make reasonable 
extrapolations of the population and per-capita GDP through `r target_year`. 

<!--
Repeat the steps below for each country or region:
-->

### Outline:

To analyze the policy:

1. Get the Kaya identity data for the OECD
2. Figure out appropriate starting years for calculating the historical trends 
   for the Kaya variables _P_, _g_, _e_, and _f_.
3. Calculate the _historical trends_ for the Kaya variables from the starting
   year you determined in step (2).
4. Use the _historical trends_ to extrapolate projected values for _P_, _g_, 
   _e_, and _f_ in `r target_year`.
5. Calculate the policy goal for emissions _F_ in `r target_year`.
   This uses the policy criteria (_F_ in `r target_year` should be
   `r label_percent(accuracy=1)(target_reduction)` less than in `r ref_year`) 
   and the measured emissions _F_ in `r ref_year`, from the Kaya data for the 
   OECD.
6. Calculate the _implied rate of change_ of _F_ between `r current_year`
   and `r target_year`, in order to reduce emissions to the policy goal that
   you calculated in step (5).
   This is the average fraction that _F_ must be reduced each year between 
   now and `r target_year` in order to meet the goal, so if the implied
   rate is $-0.05$, then _F_ must be reduced by 5% each year, on average.
7. Combine the _implied rate of change_ of _F_ with the _historical trends_ of
   _P_ and _g_ to calculate the _implied rate of change_ of _ef_ that you 
   calculated in step (3) in order to meet the policy goal from step (5).
8. Compare the _implied rate of change_ of _ef_ that you calculated in step (7)
   to the _historical trend_ of _ef_ that you can determine from the
   _historical trends_ of _e_ and _f_ that you calculated in step (3).

### Detailed steps:

1. Use the `kayadata` package in RStudio to load the data for
   the OECD. Below is an example of looking up the data
   for the United States:
   ```{r get_oecd_data, echo=TRUE, eval=TRUE, include=TRUE, warning=FALSE, message=FALSE}
   library(kayadata)
   us_data = get_kaya_data("United States")
   us_latest = us_data %>% slice_max(year)
   ```
   You can get a list of all the countries and regions that are available 
   from `kaya_region_list()`.

2. Next, open up the interactive application at 
   [ees3310.jgilligan.org/decarbonization](https://ees3310.jgilligan.org/decarbonization),
   seelect OECD as the region. 
   Leave the "Calculate trends starting in" at its default value (1980).
   Go to the "Trends" tab and look at the graphs of 
   ln(_P_), ln(_g_), ln(_e_), ln(_f_), and ln(_ef_).
   
   * For each graph compare the real data (in red) to the trend line 
     (the straight blue line).
   * Does the trend line look a like a good description of the data?
   * Is there a better starting year for calculating trends? 
     If so, adjust ``Calculate trends starting in'' to this year
     and make note of the year.
   * You should also plot these in your report using RMarkdown. 
     Following from the example above, you can use the `plot_kaya` function:
     ```{r plot_us_data, echo=TRUE, eval=TRUE, include=TRUE, warning=FALSE, message=FALSE}
     plot_kaya(us_data, "e", log_scale = TRUE, font_size = 12)
     ```
     Use plots like the one above to examine trends in _P_, _g_, _e_, and _f_.
     Be sure to set `log_scale = TRUE` in the `plot_kaya` function because 
     a constant percentage rate of
     change corresponds to a linear trend in the logarithm of the variable.
3. In your report, discuss whether you anticipate a problem if we make policy 
   by assuming that the Kaya identity variables will follow the trend line for 
   the next  several decades?

4. Next, calculate the rates of change of 
   _P_, _g_, _e_, and _f_ (the Population, per-capita GDP, energy intensity of 
   the economy, and carbon-intensity of the energy supply) from your starting 
   year through `r current_year`, using the `lm` function in R.

   A constant rate of change is represented by a linear relationship between
   the natural logarithm of the kaya variable and time: for the variable
   `P` (population), we would write this formula in R as `log(P)~year`.

   Here is an example of calculating the rate of change of _e_ 
   (the energy intensity of the economy) for the United States, using the 
   variable `us_data` that you calculated above:
   ```{r get_us_e_trend, echo=TRUE, include=TRUE, eval=TRUE, warning=FALSE, message=FALSE}
   # Load the broom library for organizing lm results
   library(broom)

   e_trend = us_data %>% filter(year >= 1980) %>%
     lm(log(e) ~ year, data = .) %>%
     tidy() %>% 
     filter(term == "year") %>% 
     pull(estimate)
   ```

   For more detailed explanation of the code above, see the handout
   "New Tools for Data Analysis."

   Here, we find that `e_trend` = `r round(e_trend, 4)` 
   (`r round(100 * e_trend, 2)`% per year).
    
   You can check your results against the interactive web application by
   looking at the rates of change reported on the "Trends" tab.
   Be sure to set the start year on the web app to the same values that you
   used for calculating the slopes in RMarkdown.

   These numbers are the slopes of the trend lines that you looked at in part 2.

5. Using the rates of change that you determined in part 4,
   use the formulas from the "Growth Rates and Trends" section to compute the 
   values for 
   _P_, _g_, _e_, and _f_ 
   in the year `r target_year`.
   
   Next, use the growth rates of _P_, _g_, _e_, and _f_ to calculate the 
   growth rate of the total emissions _F_. 
   Calculate the total CO~2~ emissions (_F_) from the OECD in 
   `r target_year`, assuming that emissions continue to grow at 
   historical rates. 
    
   It may be useful to define functions for frequently used 
   (e.g., see the example `growth` function in the handout on "New Tools for
   Data Analysis")

   Check your work against the bottom-up numbers in the ``Bottom-Up Analysis'' 
   table on the bottom of the left-hand pane of the web application.

6. Calculate the emissions target: Set the reference year for 
   emissions reduction to `r ref_year`, and set the target emissions
   reduction to `r label_percent(accuracy = 1)(target_reduction)`.

   Set the target year in the web app to `r target_year`;
   set the reference year to `r ref_year`;
   set the emissions reduction to 
   `r label_percent(accuracy=1)(target_reduction)`.

   How much CO~2~ (_F_) would the OECD emit in
   `r target_year` in order to meet your policy goal? 

   Let's work an example using the Middle East:
   ```{r middle_east_example, echo = TRUE, eval=TRUE, include=TRUE, warning=FALSE, message=FALSE}
   F_2005_middle_east = get_kaya_data("Middle East") %>% 
     filter(year == 2005) %>% pull(F)
   F_2005_middle_east
   ```
   ```{r middle_east_example_2, echo = TRUE, eval=TRUE, include=TRUE, warning=FALSE, message=FALSE}
   middle_east_reduction = 0.32
   F_goal_middle_east = F_2005_middle_east * (1 - middle_east_reduction)
   F_goal_middle_east
   ```

   Check this result against the interactive web application.

7. Look up what the CO~2~ emission is in `r current_year` and calculate the 
   rate of change in _F_ that would be necessary to achieve your policy target. 
   For the `r target_year`\ calculation:
   
   a. Calculate the ratio of $F_{`r target_year`}/F_{`r current_year`}$.
   b. Take the natural logarithm of this ratio 
      (in R, the natural logarithm function is `log()`; on your calculator
      it is "LN").
   c. Divide the logarithm by the number of years 
      ($`r target_year` - `r current_year`$). 
      This is the rate of change of _F_. A positive number means growth and a 
      negative number means a reduction.
   The percentage rate of change per year is 100 times this number.

   For our Middle East example:
   ```{r middle_east_example_3, echo = TRUE, eval=TRUE, include=TRUE, warning=FALSE, message=FALSE}
   F_2017_middle_east = get_kaya_data("Middle East") %>% 
     filter(year == 2017) %>% pull(F)
   r_F_middle_east = log(F_goal_middle_east / F_2017_middle_east) / 
     (2050 - 2017)
    r_F_middle_east
    ```
    so total emissions for the Middle East would need to drop by 
    `r label_percent(accuracy=0.01)(r_F_middle_east)` per year between 
    `r current_year` and `r target_year`

8. Now calculate the decarbonization rate implied by the policy goal. This is 
   the rate of reduction of _ef_, the carbon intensity of the economy. 
   $F = Pgef$, so $r_F = r_P + r_g + r_e + r_f$. Subtract the projected _r~P~_ 
   and _r~g~_ 
   (look them up in the ``Bottom up Analysis'' table) 
   from _r~F~_, which you just calculated in step~7, 
   to get the rate of change of _ef_. 
   Multiply the rate of change of _ef_ by -1 to get the rate of decarbonization
   (because negative rate of change is a positive rate of decarbonization and
   vice-versa). 
   Multiply by 100 to get the percent implied rate of decarbonization.

9. How does the implied rate of decarbonization for the OECD compare to the 
   historical rate of decarbonization (i.e., the trend in _ef_ reported in the 
   "Bottom up Analysis" table)?
