---
title: 'Lab #5 Answers'
lab_number: 5
lab_date: '2022-02-21'
sol_date: '2022-03-13'
pubdate: '2022-03-13'
date: '2022-03-02'
pdf_url: /files/lab_solutions/lab_05/lab_05_answers.pdf
slug: lab_05_answers
output:
  blogdown::html_page:
    md_extensions: +tex_math_single_backslash+compact_definition_lists
    toc: yes
---
```{r setup, include=FALSE}
knitr::knit_hooks$set(inline = function(x) { knitr:::format_sci(x, 'md')})
knitr::opts_chunk$set(echo = TRUE)

if (knitr::is_latex_output()) {
  message("Setting PDF plot hook.")
  new_plot_hook <- function(x, options)
    paste("\n", knitr::hook_plot_tex(x, options), "\n")

  knitr::knit_hooks$set(plot = new_plot_hook)
} else {
  message("Not using PDF output.")
}


options(dplyr.summarise.inform = FALSE)

# This section loads necessary R libraries and sources scripts that define
# useful functions format_md.
#
data_dir = "_data"
script_dir = "_scripts"

# library(pacman)
# p_load(tidyverse, jsonlite, httr, magrittr, scales)
library(tidyverse)
library(magrittr)
library(scales)
library(jsonlite)
library(xml2)
library(httr)
library(knitr)

theme_set(theme_bw(base_size = 10))

source(file.path(script_dir, "utils.R"), chdir = T)
source(file.path(script_dir, "format_md.R"), chdir = T)
source(file.path(script_dir, "modtran.R"), chdir = T)
source(file.path(script_dir, "rrtm.R"), chdir = T)
```
In the answers below, I illustrate how you can use RMarkdown to integrate 
code and data analysis with your text to produce a clear report that follows the
principles of reproducible research

# Exercise 1: Clouds and Infrared.

Clouds play an important and complex role in the climate system because they
interact with both visible and infrared light. In this exercise, I investigated
the way that the clouds interact with the flow of heat in the form of longwave
infrared light.

I began by comparing the longwave radiation going out from the atmosphere to
space under four conditions: with no clouds and with low, medium, and 
high-altitude clouds. Next, I compared the upward longwave radiation going out
to space with the radiation going downward from the atmosphere, which is 
absorbed by the surface, and compared the upward and downward radiation for 
clear skies and with low clouds. Finally, I investigated the effect of water
vapor on the downward radiation, which let me compare the impact of water vapor 
to the impact of clouds.

The earth's temperature is determined by balancing incoming and outgoing 
radiation. Incoming radiation reaches the earth as shortwave radiation from the
sun and for this lab, I will only look at the effect clouds have on longwave
radiation, so I will ignore the way clouds affect shortwave radiation (i.e., 
I will ignore the way that clouds change the earth's albedo).

## Clouds and Outgoing Longwave Radiation

First, I look at the effect clouds have on outgoing radiation. 
To do this, I start by running MODTRAN with the default settings, which include
a clear sky with no clouds:
```{r modtran_default, fig.cap = "Looking down from 70 km with a clear sky."}
default_outgoing = run_modtran(atmosphere = "tropical", altitude_km = 70, 
                               looking = "down")
i_out_clear = default_outgoing$i_out
plot_modtran(default_outgoing, descr = "Clear sky, looking down from 70 km")
```

Next, I do the same for three kinds of clouds: low-altitude stratus clouds,
medium altitude altostratus, and high-altitude cirrus.

```{r clouds-looking-down}
stratus_outgoing = run_modtran(atmosphere = "tropical", altitude_km = 70, 
                               looking = "down", clouds = "stratus")
altostratus_outgoing = run_modtran(atmosphere = "tropical", altitude_km = 70, 
                               looking = "down", clouds = "altostratus")
cirrus_outgoing = run_modtran(atmosphere = "tropical", altitude_km = 70, 
                               looking = "down", clouds = "standard cirrus")

i_out_stratus = stratus_outgoing$i_out
i_out_altostratus = altostratus_outgoing$i_out
i_out_cirrus = cirrus_outgoing$i_out
```
```{r stratus-down-fig, fig.cap = "Looking down on stratus clouds from 70 km"}
plot_modtran(stratus_outgoing, 
                         descr = "Stratus clouds, looking down from 70 km")
```
```{r altostratus-down-fig, fig.cap = "Looking down on altostratus clouds from 70 km"}
plot_modtran(altostratus_outgoing, 
                         descr = "Altostratus clouds, looking down from 70 km")
```
```{r cirrus-down-fig, fig.cap = "Looking down on cirrus clouds from 70 km"}
plot_modtran(cirrus_outgoing, 
                         descr = "Cirrus clouds, looking down from 70 km")
```


The following table compares the total intensity of outgoing radiation with 
different kinds of clouds and with a clear sky:
```{r i-out-table}
i_out_tbl = tibble(clouds = c("none", "stratus", "altostratus", "cirrus"),
                   altitude = c("", "low", "medium", "high"),
                   i_out = c(i_out_clear, i_out_stratus, i_out_altostratus,
                             i_out_cirrus),
                   change = i_out - i_out_clear
                   )
kable(i_out_tbl, digits = 1)
```

Adding stratus (low-altitude) clouds reduces I~out~ by 
`r format_md(i_out_clear - i_out_stratus, digits = 2)` W/m^2^,
adding cirrus (high-altitude) clouds reduces I~out~ by 
`r format_md(i_out_clear - i_out_cirrus, digits = 2) `W/m^2^, and
altostratus (medium-altitude) clouds reduces I~out~ by 
`r format_md(i_out_clear - i_out_altostratus, digits = 2) `W/m^2^.

When we compare the spectra of outgoing radiation, there is very little 
difference in the parts of the spectrum where greenhouse gases absorb 
strongly (CO~2~ near 700 cm^-1^, ozone near 1050 cm^-1^,
and water vapor in the regions below about 550 cm^-1^ and above 1200 cm^-1^),
but there are clear differences in the two atmospheric window regions
(800--1000 cm^-1^ and 1100--1200 cm^-1^), where the greenhouse gases do not 
absorb very much.
For all three cloud types, the outgoing radiation is smaller in both window 
regions. The difference is greatest for altostratus, less for cirrus, and 
least of all for stratus.


What is happening here is the radiation we see with clear skies in the window
region is coming from the very low atmosphere, near the surface. When there are
clouds, they are higher up, so they block the radiation from near the surface,
but they also emit radiation at an intensity that corresponds to the temperature
of the tops of the clouds.

The higher up you go, the colder the temperature is, so the cold tops of the 
clouds emit longwave radiation with lower intensity than the atmosphere below
them would.

The difference between stratus and altostratus has to do with the altitude of
the clouds: higher altitudes correspond to lower temperatures, so the tops of 
altostratus clouds (medium altitude) are colder than the tops of the stratus
clouds (low-altitude).

With cirrus, the situation is more complicated. Stratus and altostratus clouds
tend to cover the whole sky, but cirrus clouds are wispy and only cover a small
fraction of the sky (see the pictures from my lecture for class #8 on
Feb. 10: <https://ees3310.jgilligan.org/slides/class_08/#/cloud-feedbacks-2>
and <https://ees3310.jgilligan.org/slides/class_08/#/cloud-feedbacks-3>).
This means that the cirrus clouds will have a smaller effect on outgoing 
radiation than the altostratus clouds, even though their temperature is
lower. 
(**Note:** Students do not need to get the explanation of cirrus clouds correct 
to get full credit for this exercise.)

## Looking up from the Ground with a Clear Sky

We have spent a lot of time looking down from high in the atmosphere, but 
MODTRAN also lets us look up from the ground and see the longwave light coming
down to the surface from the atmosphere. We start by comparing the spectrum of 
longwave radiation going up to space and going down to the surface under a 
clear sky:

```{r clear-up, fig.cap = "Upward longwave radiation going out to space from a clear sky"}
clear_up = run_modtran(atmosphere = "midlatitude winter", 
                         altitude_km = 70, looking = "down",
                         h2o_scale = 1, clouds = "none")
plot_modtran(clear_up, descr = "Looking down from 70 km with clear skies")
```


```{r clear-down, fig.cap = "Downward longwave radiation coming to the surface from a clear sky"}
clear_down = run_modtran(atmosphere = "midlatitude winter", 
                         altitude_km = 0, looking = "up",
                         h2o_scale = 1, clouds = "none")
i_down_clear = clear_down$i_out
plot_modtran(clear_down, descr = "Looking up from 0 km with clear skies")
```

When we compare the two spectra, they are almost opposite. The upward radiation 
going out to space is brightest in the atmospheric window and dimmest where 
CO~2~ absorbs strongly.
The downward radiation reaching the ground is brightest where CO~2~ and water 
vapor absorb strongly and dimmest in the atmospheric window.

The reason for this is that where the atmosphere is most transparent, a lot of 
the radiation we see is coming from whatever is on the other side of the 
atmosphere. When we are on the surface looking up with our eyes, and we aren't 
looking at the sun we see visible light from space (the moon and stars).
Outer space is very cold, so in the atmospheric window region of the longwave 
spectrum, our sensor looking up from the surface sees blackbody radiation 
corresponding to the cold temperatures of space, and our sensor looking down
from 70 km sees blackbody radiation corresponding to the warm temperature of the
surface.

Where CO~2~ absorbs strongly, when we look down from space, the CO~2~ in the 
atmosphere absorbs the bright radiation from the warm lower atmosphere before it
can get up to the sensor, so what we see is the blackbody radiation from CO~2~
molecules high up in the atmosphere where the temperature is cold, so the 
radiation is very faint. When we look up from the surface, CO~2~ molecules 
near the surface have high emissivity, so they radiate strongly and the sensor
sees intense radiation. The same applies to water vapor, and the sensor sees 
bright, intense radiation in the parts of the spectrum where water vapor absorbs
strongly (i.e., where it has a large emissivity).

## Looking up from the Ground with Clouds

Next, I investigated how the presence of clouds changes the spectrum that a 
sensor looking up from the ground will see.

```{r stratus-down, fig.cap = "Downward longwave radiation coming to the surface from stratus clouds"}
stratus_down = run_modtran(atmosphere = "midlatitude winter", 
                         altitude_km = 0, looking = "up",
                         h2o_scale = 1, clouds = "stratus")
i_down_stratus = stratus_down$i_out
plot_modtran(stratus_down, descr = "Looking up from 0 km with stratus clouds")
```

Here, the spectrum looks like a perfect black body at a temperature of 
around 270 K.
The bottom of the cloud acts like a perfect blackbody, and stratus clouds are
low in the sky, so the temperature is relatively high. This has an especially 
big effect in the atmospheric window regions of the spectrum. In the windows,
the downward spectrum was very faint with clear skies, but with stratus clouds
the windows have a lot of downward radiation.

Basically, the bottoms of the low clouds are warm, and because of this, they 
glow brightly with longwave blackbody radiation, and that sends a lot of heat to 
the surface. Adding stratus clouds to a winter sky increased the heat coming to
the surface from the atmosphere from 
`r format_md(i_down_clear, digits = 3)` Watts per square meter under clear
skies to `r format_md(i_down_stratus, digits = 3)` Watts per square meter.

This extra heat will warm the ground, so on a winter night, we expect the 
temperature to be a good deal warmer when the sky is cloudy than when it is
clear.

## The Effect of Water Vapor

Next, I investigated the way water vapor affects outgoing longwave radiation and
compare that to the effects of clouds. To do this, I compare the intensity of
downward longwave radiation at the surface with the normal amount of water vapor
in the atmosphere to the intensity when there is no water vapor.

The previous clear-sky run had a normal amount of water vapor, so I only need to
run MODTRAN with `h2o_scale = 0` and compare that to the previous `clear_down`
run, which had the default value `h2o_scale = 1`

```{r dry-down, fig.cap = "Downward longwave radiation coming to the surface with a clear sky and no water vapor"}
dry_down = run_modtran(atmosphere = "midlatitude winter", 
                         altitude_km = 0, looking = "up",
                         h2o_scale = 0, clouds = "none")
i_down_dry = dry_down$i_out
plot_modtran(dry_down, 
             descr = "Looking up from 0 km with a clear sky and no water vapor")
```

This spectrum contrasts dramatically with the spectrum of downward radiation
for clear skies with a normal amount of water vapor. In the regions below
about 550 cm^-1^ and above 1200 cm^-1^, the longwave radiation is much less 
intense---indeed, there is basically none, except for the region around 1300 
cm^-1^ where methane absorbs and emits.

The three big features of longwave emission correspond to the CO~2~ absorption
near 700 cm^-1^, ozone absorption near 1050 cm^-1^, and methane absorption near
1300 cm^-1^. Other than those three features, there is essentially no downward
radiation. This is because without any water vapor, the atmosphere is mostly
transparent to longwave radiation, and because good absorbers are also good 
emitters, this means that the amosphere emits almost no radiation in most
of the spectrum. What the sensor sees from the ground in most of the spectrum
is just the cold temperature of space, and the correspondingly low intensity of
longwave radiation.

The total downward radiation with clear skies and no water vapor is
`r format_md(i_down_dry, digits = 2)` W/m^2^, which is 
`r format_md(i_down_clear - i_down_dry, digits = 3)` W/m^2^ less than for clear
skies with water vapor. This difference is a way to measure the greenhouse effect
of water vapor: adding water vapor to the atmosphere increases the amount of
heat absorbed by the surface by 
`r format_md(i_down_clear - i_down_dry, digits = 3)` W/m^2^. 
This heat is longwave radiation emitted by the water vapor in the atmosphere, 
and it heats up the earth's surface 
(the ground and the water at the surface of the oceans).

Water vapor has a much larger effect than adding clouds. 
Adding stratus clouds to a clear sky increased I~out~ by 
`r format_md(i_down_stratus - i_down_clear, digits = 2)` W/m^2^
and adding water vapor to a dry atmosphere increased I~out~ by
`r format_md(i_down_clear - i_down_dry, digits = 3)` W/m^2^.

# Exercise 2: Water Vapor Feedback

In the previous exercise I found that water vapor affects longwave radiation
very powerfully. The amount of water vapor in the atmosphere depends on the
air temperature. The warmer the air is, the more water vapor it can hold, so
as temperatures rise, the amount of water vapor also rises, and as temperatures
fall, the amount of water vapor falls. This effect creates a powerful 
positive feedback involving water vapor.

Here I will estimate the effect of the water-vapor feedback using the RRTM
climate model, which includes the effects of both radiation and convection.
To measure the water vapor feedback, I will measure the climate sensitivity
under normal conditions and when there is no water vapor in the atmosphere.

## Finding the Climate Sensitivity

The climate sensitivity is defined as the change in surface temperature that
occurs when the amount of CO~2~ is doubled. To calculate the climate sensitivity
with the RRRTM model, I first ran the model with CO~2~ at its current value of 
440 ppm, and make sure that the heat flow into and out of the earth is balanced.
Then I ran the model with CO~2~ doubled to 800 ppm and adjusted the surface 
temperature until the model reported that the heat flow was balanced.

First, I ran the RRTM model with its default parameters.

```{r rrtm-default}
rrtm_default = run_rrtm(co2_ppm = 400, relative_humidity = 80)
q_default = rrtm_default$Q
```

The results from RRTM model runs include the variable _Q_, which is the 
imbalance of I~in~ - I~out~, so if _Q_ > 0, there is more heat coming in and
the earth will warm up, and if _Q_ < 0, there is more heat going out and the
earth will cool off.

Here, I can verify that the RRTM model in its default configuration has 
_Q_ = `r format_md(q_default, digits = 1)` W/m^2^.

Next, I ran RRTM with CO~2~ doubled:

```{r rrtm-doubled}
rrtm_doubled = run_rrtm(co2_ppm = 800, relative_humidity = 80)
q_doubled = rrtm_doubled$Q
```

With CO~2~ doubled, _Q_ becomes `r format_md(q_doubled, digits = 1)` W/m^2^.
Using trial and error with the interactive web version of RRTM, I found a 
temperature where _Q_ = 0.
Initially, _Q_ was `r format_md(q_doubled, digits = 1)`, which is 
`r ifelse(q_doubled > 0, "greater", "less")` than 0, so I knew that I had to 
`r ifelse(q_doubled > 0, "increase", "decrease")` the surface temperature.
I first raised the surface temperature from 
`r format_md(rrtm_doubled$T_surface, digits = 3)` K to 290 K.

```{r rrtm-test-1}
test_T_1 = 290
rrtm_test_1 = run_rrtm(co2_ppm = 800, relative_humidity = 80, 
                       T_surface = test_T_1)
q_test_1 = rrtm_test_1$Q
```

Now, _Q_ = `r format_md(q_test_1, digits = 1)`, which is less than zero, so 
I needed to reduce T~surface~. I chose a value roughly halfway between the 
original temperature (`r format_md(rrtm_doubled$T_surface, digits = 3)` K) and
290 K: my new guess was 287 K.

```{r rrtm-test-2}
test_T_2 = 287
rrtm_test_2 = run_rrtm(co2_ppm = 800, relative_humidity = 80, 
                       T_surface = test_T_2)
q_test_2 = rrtm_test_2$Q
```

Now, _Q_ = `r format_md(q_test_2, digits = 1)`, which is still slightly less 
than zero, so I needed to reduce T~surface~ a bit more. 
I chose a value roughly halfway between the original temperature 
(`r format_md(rrtm_doubled$T_surface, digits = 3)` K) and `r test_T_2` K: 
my new guess was 286 K

```{r rrtm-test-3}
test_T_3 = 286
rrtm_test_3 = run_rrtm(co2_ppm = 800, relative_humidity = 80, 
                       T_surface = test_T_3)
q_test_3 = rrtm_test_3$Q
```

Now, _Q_ = `r format_md(q_test_3, digits = 1)`, so I guess a new temperature
halfway between `r test_T_3` and `r test_T_2`

```{r rrtm-test-4}
test_T_4 = 286.5
rrtm_test_4 = run_rrtm(co2_ppm = 800, relative_humidity = 80, 
                       T_surface = test_T_4)
q_test_4 = rrtm_test_4$Q
```

Now, _Q_ = `r format_md(q_test_4, digits = 1)`, so I guess a new temperature
halfway between `r test_T_4` and `r test_T_2`

```{r rrtm-test-5}
test_T_5 = 286.75
rrtm_test_5 = run_rrtm(co2_ppm = 800, relative_humidity = 80, 
                       T_surface = test_T_5)
q_test_5 = rrtm_test_5$Q
```

Now, _Q_ = `r format_md(q_test_5, digits = 1)`, so I guess a new temperature
halfway between `r test_T_5` and `r test_T_2`

```{r rrtm-test-6}
test_T_6 = 286.87
rrtm_test_6 = run_rrtm(co2_ppm = 800, relative_humidity = 80, 
                       T_surface = test_T_6)
q_test_6 = rrtm_test_6$Q
```


Now, _Q_ = `r format_md(q_test_6, digits = 1)`, so I am done and the new
equilibrium surface temperature is `r test_T_6`.
The climate sensitivity for an atmosphere with 80% average humidity is the 
change in surface temperature that was necessary to bring _Q_ to zero.

```{r t-2xco2}
climate_sensitivity_humid = test_T_6 - rrtm_doubled$T_surface
```

The climate sensitivity is 
`r format_md(climate_sensitivity_humid, digits = 2)` K

## Climate Sensitivity without Water Vapor

Next, I do the same exercise, but with `relative_humidity = 0` so there will be
no water-vapor feedback.

First, I run RRTM with no water vapor and 400 ppm CO~2~:

```{r rrtm-dry}
rrtm_dry = run_rrtm(co2_ppm = 400, relative_humidity = 0)
q_dry = rrtm_dry$Q
```

_Q_~dry~ = `r format_md(q_dry, digits = 1)`, so before I can do anything else, 
I need to adjust T~surface~ to bring _Q_ to zero.

```{r rrtm-dry-equilibrium}
new_T_surface = 261.55
rrtm_dry_eq = run_rrtm(co2_ppm = 400, relative_humidity = 0, 
                       T_surface = new_T_surface)
q_dry_eq = rrtm_dry_eq$Q
```

Repeating the same trial-and-error process illustrated above, I tried
T~surface~ = 270 K, then 260 K, then 265 K, then 262.5 K, then 261 K, then
261.5 K, then 261.6 K, then 261.55 K, which finally gave
_Q_ = `r round(q_dry_eq, 1)`.

Next, I doubled CO~2~

```{r rrtm-doubled-dry}
rrtm_doubled_dry = run_rrtm(co2_ppm = 800, relative_humidity = 0, 
                            T_surface = new_T_surface)
q_doubled_dry = rrtm_doubled_dry$Q
```


_Q_~doubled,dry~ = `r format_md(q_dry, digits = 1)`, so before I can do 
anything else, I need to adjust T~surface~ to bring _Q_ to zero.

```{r rrtm-doubled-dry-eq, eval = TRUE, include = FALSE}
climate_sensitivity_dry = 1.1
T_doubled_dry = new_T_surface + climate_sensitivity_dry
rrtm_doubled_dry_eq = run_rrtm(co2_ppm = 800, relative_humidity = 0, 
                            T_surface = T_doubled_dry)
q_doubled_dry_eq = rrtm_doubled_dry_eq$Q
```

After some trial and error, similar to what I showed above, I found that
raising the surface temperature by `r round(climate_sensitivity_dry, 1)` K 
would bring _Q_ to `r round(q_doubled_dry_eq, 2)`

```{r rrtm-doubled-dry-eq, echo = TRUE, eval = FALSE}
```

## Assessing the Water-Vapor Feedback

Without any water vapor in the atmosphere, the climate sensitivity is
`r format_md(climate_sensitivity_dry, digits = 2)` K 
and with water vapor in the atmosphere and the water-vapor feedback operating,
the climate sensitivity rises to 
`r format_md(climate_sensitivity_humid, digits = 2)` K, so the 
water vapor feedback amplifies the climate sensitivity by a factor of 
`r format_md(climate_sensitivity_humid / climate_sensitivity_dry, digits = 1)`.

This means that when the amount of CO~2~ in the atmosphere rises, 
the water-vapor feedback more than doubles the amount of warming.
