---
title: 'Lab #2 Answers'
lab_number: 2
lab_date: '2020-01-13'
sol_date: '2020-02-03'
pubdate: '1580709600'
date: '2020-01-24'
pdf_url: /files/lab_solutions/lab_02/lab_02_answers.pdf
slug: lab_02_answers
output:
  blogdown::html_page:
    md_extensions: +tex_math_single_backslash+compact_definition_lists
    toc: yes
  pdf_document:
    md_extensions: +tex_math_single_backslash+compact_definition_lists
    toc: yes
    includes:
      in_header: ees3310.sty
---
```{r setup, include=FALSE}
knitr::knit_hooks$set(inline = function(x) { knitr:::format_sci(x, 'md')})
knitr::opts_chunk$set(echo = TRUE)

# This section loads necessary R libraries and sources scripts that define 
# useful functions format_md.
# 
data_dir = "_data"
script_dir = "_scripts"

library(pacman)
p_load(tidyverse, zoo)

theme_set(theme_bw(base_size = 15))

source('_scripts/utils.R', chdir = T)
source('_scripts/modtran.R', chdir = T)

mlo_url = str_c("https://scrippsco2.ucsd.edu/assets/data/atmospheric/stations/",
                "in_situ_co2/monthly/monthly_in_situ_co2_mlo.csv")

giss_url = "https://data.giss.nasa.gov/gistemp/tabledata_v4/GLB.Ts+dSST.csv"

# Create a data directory if one does not exist.
if (!dir.exists('_data')) dir.create('_data')
```
```{r download_mlo_data, echo = FALSE, include=FALSE, message=FALSE}
download.file(mlo_url, '_data/mlo_data.csv')
```
```{r import_mlo_data, echo=FALSE, include=FALSE, message=FALSE}
mlo_data = read_csv('_data/mlo_data.csv', 
                    skip = 57,  # skip the first 57 rows
                    col_names = c('year', 'month', 'date.excel', 'date',
                                  'co2.raw', 'co2.raw.seas', 
                                  'co2.fit', 'co2.fit.seas',
                                  'co2.filled', 'co2.filled.seas'),
                    col_types = 'iiiddddddd', 
                    # ^^^ the first three columns are integers and the  next 
                    # 7 are real numbers
                    na = '-99.99' 
                    # ^^^ interpret -99.99 as a missing value
)
```
```{r show_data_r, echo=FALSE, include=FALSE}
head(mlo_data)
```
```{r simplify_mlo_data, echo=FALSE, include=FALSE}
mlo_simple = mlo_data %>% select(year, month, date, co2 = co2.filled)

head(mlo_simple)
```
```{r show_data, echo=FALSE, include=FALSE, results = "asis"}
head(mlo_simple) %>% 
  knitr::kable(col.names = c(year = "Year", month = "Month", date = "Date",
                             co2 = "CO~2~ (ppm)"),
               caption = "A table of monthly CO~2~ measurements from Mauna Loa.")
```
```{r plot_mlo, echo=FALSE, include=FALSE, warning = FALSE, fig.show = "hold", fig.cap="Monthly CO~2~ measurements from Mauna Loa."}
ggplot(mlo_simple, aes(x = date, y = co2)) + 
  # ^^^ The ggplot command specifies which data to plot and the aesthetics that 
  # define which variables to  use for the x and y axes.
  geom_line() +   
  # ^^^ The geom_line() command says to plot lines connnecting the data points
  labs(x = "Year", y = "CO2 concentration (ppm)",
       title = "Measured CO2 from Mauna Loa Observatory")
  # ^^^ The labs() command tells ggplot what names to use in labeling the axes 
  # and the title for the plot.

  # Earlier in this .Rmd file, I called set_theme(theme_bw(base_size = 15)) 
  # to set the default plot style. If you call ggplot() without this,
  # you will get a different style, but you can either call theme_set
  # or you can add a theme specification (such as 
  # "+ theme_bw(base_size = 15)")
  # to the end of the sequence of plotting commands in order to 
  # apply a specific style to an individual plot.
```
```{r plot_mlo_annual, echo=FALSE, include=FALSE, warning = FALSE, fig.cap="Raw and seasonally adjusted measurements of atmospheric CO~2~, from Mauna Loa."}
mlo_simple %>% mutate(annual = rollapply(data = co2, width = 12, FUN = mean,
                                      fill = NA, align = "center")) %>%
  ggplot(aes(x = date)) + 
  geom_line(aes(y = co2), color = "darkblue", size = 0.1) +
  geom_line(aes(y = annual), color = "black", size = 0.5) +
  labs(x = "Year", y = "CO2 concentration (ppm)", 
       title = "Measured and Seasonally Adjusted CO2")
```
```{r plot_mlo_annual_2, echo=FALSE, include=FALSE, warning=FALSE, , fig.cap="Raw and seasonally adjusted measurements of atmospheric CO~2~, from Mauna Loa, with a legend identifying the different lines."}
mlo_simple %>% mutate(annual = rollapply(data = co2, width = 12, FUN = mean,
                                      fill = NA, align = "center")) %>%
  ggplot(aes(x = date)) + 
  geom_line(aes(y = co2, color = "Raw"), size = 0.1) +
  geom_line(aes(y = annual, color = "12-month average"), size = 0.5) +
  scale_color_manual(values = c("Raw" = "darkblue", 
                                "12-month average" = "black"),
                     name = "Smoothing") +
  labs(x = "Year", y = "CO2 concentration (ppm)", 
       title = "Measured and Seasonally Adjusted CO2")
```
```{r calc_mlo_trend, echo=FALSE, include=FALSE}
trend = lm(co2 ~ date, data = mlo_simple)

library(broom)

tidy(trend)
```
```{r plot_mlo_with_fitted_trend, echo=FALSE, include=FALSE, warning=FALSE, fig.cap="Trend in atmospheric CO~2~."}
mlo_simple %>% 
  ggplot(aes(x = date, y = co2)) + 
  geom_line() +
  geom_smooth(method = 'lm') +
  labs(x = "Year", y = "CO2 concentration (ppm)", 
       title = "Measured CO2 and Linear Fit")
```
# Exercises

## Exercises with CO~2~ Data from the Mauna Loa Observatory

Using the `select` function, make a new data tibble called `mlo_seas`, from 
the original `mlo_data`, which only has two columns: `date` and 
`co2.seas`, where `co2.seas` is a renamed version of `co2.filled.seas` from the 
original tibble.

```{r make_mlo_seas, include=TRUE}
# We only need to load the libraries once and they will be loaded for all 
# subsequent code chunks
library(tidyverse)
library(zoo)

mlo_seas = select(mlo_data, date, co2.filled.seas)
mlo_seas = rename(mlo_seas, co2.seas = co2.filled.seas)

# Alternately, you can simplify with the pipe operator:
# 
# mlo_seas = select(mlo_data, date, co2.filled.seas) %>% rename(co2.seas = co2.filled.seas)
# 
# or you can rename as part of the select operation:
# 
# mlo_seas = select(mlo_data, date, co2.seas = co2.filled.seas)

# Display the first few rows:
head(mlo_seas)
```

Now plot this with `co2.seas` on the _y_ axis and `date` on the _x_ axis,
and a linear fit:

```{r plot_mlo_seas, include = TRUE}
ggplot(mlo_seas, aes(x = date, y = co2.seas)) +
  geom_line() +
  geom_smooth(method="lm") +
  labs(x = "Year", y = "CO2 concentration (ppm)")
```

Now fit a linear function to find the annual trend of `co2.seas`. Save
the results of your fit in a variable called `trend.seas`.

```{r fit_mlo_sease, include=TRUE}
trend.seas = lm(co2.seas ~ date, data = mlo_seas)

tidy(trend.seas)
```

Compare the trend you fit to the raw `co2.filled` data to the trend you fit
to the seasonally adjusted data.

**Note:** I just intend students to informally look at the trend in the graph 
and estimate its slope by eye to compare to the results in `trend.seas`.

## Exercises with Global Temperature Data from NASA

We can also download a data set from NASA's Goddard Institute for Space Studies
(GISS), which contains the average global temperature from 1880 through the 
present.

The URL for the data file is 
<`r giss_url`>

Download this file and save it in the directory `_data/global_temp_land_sea.csv`.

```{r download_giss_temp, include=TRUE}
download.file(giss_url, file.path(data_dir, "global_temp_land_sea.csv"))
```

* Open the file in Excel or a text editor and look at it.

* Unlike the CO~2~ data file, this one has a single line with the 
data column names, so you can specify `col_names=TRUE` in `read_csv`
instead of having to write the column names manually.

* How many lines do you have to tell `read_csv` to skip?

**Answer:** 1 line: the first line is "Land-Ocean: Global Means" and we want to 
skip it.

* `read_csv` can automatically figure out the data types for each column,
  so you don't have to specify `col_types` when you call `read_csv`

* This file uses `***` to indicate missing values instead of `-99.99`, so you
  will need to specify `na="***"` in `read_csv`.  
  
    For future reference,
    if you have a file that uses multiple different values to indicate missing
    values, you can give a vector of values to `na` in `read_csv`:
    `na = c("***","-99.99", "NA", "")` would tell `read_csv` that if it finds 
    any of the values "***", "-99.99", "NA", or just a blank with nothing in it,
    any of those would correspond to a missing value, and should be indicated by
    `NA` in R.
  
Now read the file into R, using the `read_csv` function, and assign
the resulting tibble to a variable `giss_temp`

```{r read_giss_temp, include=TRUE}
giss_temp = read_csv(file.path(data_dir, "global_temp_land_sea.csv"), 
                     skip = 1, na = "***", col_names = TRUE)

# show the first 5 lines of giss_temp
head(giss_temp, 5)
```

Something is funny here: Each row corresponds to a year, but there are columns
for each month, and some extra columns called "J-D", "D-N", "DJF", "MAM", "JJA",
and "SON". These stand for average values for the year from January through 
December, the year from the previous December through November, and the seasonal
averages for Winter (December, January, and February), 
Spring (March, April, and May), Summer (June, July, and August), and Fall 
(September, October, and November).

The temperatures are recorded not as the thermometer reading, but as _anomalies_.
If we want to compare how temperatures are changing in different seasons and at
different parts of the world, raw temperature measurements are hard to work with
because summer is hotter than winter and Texas is hotter than Alaska, so it 
becomes difficult to compare temperatures in August to temperatures in January,
or temperatures in Texas to temperatures in Alaska
and tell whether there was warming.

To make it easier and more reliable to compare temperatures at different times
and places, we define anomalies: The temperature anomaly is the difference between
the temperature recorded at a certain location during a certain month and
a baseline reference value, which is the average temperature for that month
and location over a period that is typically 30 years.

The GISS temperature data uses a baseline reference period of 1951--1980, so 
for instance, the temperature anomaly for Nashville in July 2017 would be
the monthly average temperature measured in Nashville during July 2017 minus
the average of all July temperatures measured in Nashville from 1951--1980.

The GISS temperature data file then averages the temperature anomalies over all
the temperature-measuring stations around the world and reports a global average
anomaly for every month from January 1880 through the latest measurements
available (currently, November 2019).

Let's focus on the months only. Use `select` to select just the columns for 
"Year" and January through December (if you are selecting a consecutive range
of columns between "Foo" and "Bar", you can call `select(Foo:Bar)`).
Save the result in a variable called `giss_monthly`

```{r make_giss_monthly, include=TRUE}
giss_monthly = select(giss_temp, Year:Dec)
#
# alternately, you could remove unwanted columns:
# 
# giss_monthly = select(giss_temp, -(`J-D`:SON))
# 
# You have to use back-quotes for the column `J-D` because its name includes
# characters other than "a"-"z", "A"-"Z", "0"-"9", ".", and "_".
# You can give columns names with other characters than these, but it becomes
# more complicated to indicate them to R.

head(giss_monthly)
```

Next, it will be difficult to plot all of the data if the months are organized as
columns. What we want is to transform the data tibble into one with three columns:
"year", "month", and "anomaly". We can do this easily using the `gather` function
from the `tidyverse` package: `gather(df, key = month, value = anomaly, -Year)`
or `df %>% gather(key = month, value = anomaly, -Year)` will gather all of the 
columns except `Year` (the minus sign in `select` or `gather` means to include
all columns except the ones indicated with a minus sign) and:

* Make a new tibble with three columns: "Year", "month", and "anomaly"
* For each row in the original tibble, make rows in the new tibble for each of 
  the columns "Jan" through "Dec", putting the name of the column in "month" 
  and the anomaly in "anomaly".

Here is an example of using `gather`, using the built-in data set `presidents`,
which lists the quarterly approval ratings for U.S. presidents from 1945--1974:

```{r gather_example_prep, include=TRUE}
df = presidents@.Data %>% matrix(ncol=4, byrow = TRUE) %>%
  as_tibble() %>% set_names(paste0("Q", 1:4)) %>% mutate(year = 1944 + seq(n()))


print("First 10 rows of df are")
print(head(df, 10))
```

For each year, the table has a column for the year and 
four columns (Q1 ... Q4) that hold the quarterly 
approval ratings for the president in that quarter.
Now we want to gather these data into three columns:
one column for the year, one column to indicate the quarter,
and one column to indicate the approval rating.

We do this with the `gather` function from the `tidyverse` package.

```{r gather_example, include=TRUE}
dfg <- df %>% gather(key = quarter, # create a column called "quarter" to store
                                    # the names of the columns that are gathered
                     value = approval, # create a column called "approval" to
                                       # store the values from those columns
                                       # (i.e., the approval ratings in that 
                                       # quarter)
                     -year # the minus sign means gather all columns EXCEPT year.
                     ) %>%
  arrange(year, quarter) # sort the rows of the resulting tibble to put
                         # the years in ascending order, from 1945 to 1971
                         # and within each year, sort the quarters from Q1
                         # to Q4

head(dfg) # print the first few rows of the tibble.
```

Now you try to do the same thing to:

* First select just the columns of `giss_monthly` for the year and the 
  individual months.

* Next, gather all the months togeher, so there will be three columns:
  one for the year, one for the name of the month, and one for the 
  temperature anomaly in that month.

* Store the result in a new variable called `giss_g`

```{r gather_giss, include=TRUE}
giss_g = gather(giss_monthly, key = month, value = anomaly, -Year)
```

Remember how the CO~2~ data had a column `date` that had a year plus a fraction
that corresponded to the month, so June 1960 was 1960.4548?

Here is a trick that lets us do the same for the `giss_g` data set.
R has a data type called `factor` that it uses for managing categorical data,
such as male versus female, Democrat versus Republican, and so on.
Categorical factors have a textual label, but are silently represented as integer 
numbers. Normal factors don't have a special order, so R sorts the values alphabetically.
However, there is another kind of factor called an ordered factor, which allows
us to specify the order of the values.

We can use a built-in R variable called `month.abb`, which is a vector of
abbreviations for months.

The following command will convert the `month`  column in `giss_g` into an
ordered factor that uses the integer values 1, 2, ..., 12 to stand for
"Jan", "Feb", ..., "Dec", and then uses those integer values to create a new
column, `date` that holds the fractional year, just as the `date` column in
`mlo_data` did:

```
giss_g = giss_g %>% 
  mutate(month = ordered(month, levels = month.abb),
         date = Year + (as.integer(month) - 0.5) / 12) %>% 
  arrange(date)`
```

In the code above, `ordered(month, levels = month.abb)` converts the variable
`month` from a character (text) variable that contains the name of the month
to an ordered factor that associates a number with each month name, such that
"Jan" = 1 and "Dec" = 12.

Then we create a new column called `date` to get the fractional year corresponding
to that month. We have to explicitly convert the ordered factor into a number
using the function `as.integer()`, and we subtract 0.5 because the time that
corresponds to the average temperature for the month is the middle of the month.

Below, use code similar to what I put above to add a new `date` column to
`giss_g`.

```{r add_date_to_giss_g, include=TRUE}
# Here, you just copy the code from above and run it.
# 
giss_g = giss_g %>% 
  mutate(month = ordered(month, levels = month.abb),
         date = Year + (as.integer(month) - 0.5) / 12) %>%
  arrange(date)
```

Now plot the monthly temperature anomalies versus date:

```{r plot_giss, include=TRUE}
ggplot(giss_g, aes(x = date, y = anomaly)) +
  geom_line() +
  geom_point() +
  labs(x = "Year", y = "Temperature Anomaly (Celsius)")
```

That plot probably doesn't look like much, because it's very noisy.
Use the function `rollapply` from the package `zoo` to create 
new columns in `giss_g` with
12-month and 10-year (i.e., 120-month) rolling averages of the 
anomalies.

Make a new plot in which you plot a thin blue line for the monthly anomaly
(use `geom_line(aes(y = anomaly), color = "blue", alpha = 0.3, size = 0.1)`;
alpha is an optional specification for transparency where 0 means invisible
(completely transparent) and 1 means opaque),
a medium dark green line for the one-year rolling average,
and a thick dark blue line for the ten-year rolling average.

```{r plot_giss_with_smoothing, include=TRUE, warning=FALSE}
giss_g %>% 
   mutate( smooth.1 = rollapply(data = anomaly, width = 12, FUN = mean,
                                      fill = NA, align = "center"),
           smooth.10 = rollapply(data = anomaly, width = 120, FUN = mean,
                                      fill = NA, align = "center")) %>% 
  ggplot(aes(x = date)) + # Put code here to map variables to aesthetics
  geom_line(aes(y = anomaly), alpha = 0.3, size = 0.1, color = "blue") +
  geom_line(aes(y = smooth.1), color = "green", size = 0.5) + 
  geom_line(aes(y = smooth.10), color = "blue", size = 1) +
  labs(x = "Year", y = "Temperature Anomaly (Celsius)")
```

Alternately, we could do this fancier version:
```{r alt_plot_giss_with_smoothing, include=TRUE, warning=FALSE}
giss_g %>% 
   mutate( smooth.1 = rollapply(data = anomaly, width = 12, FUN = mean,
                                      fill = NA, align = "center"),
           smooth.10 = rollapply(data = anomaly, width = 120, FUN = mean,
                                      fill = NA, align = "center")) %>% 
  ggplot(aes(x = date)) + # Put code here to map variables to aesthetics
  geom_line(aes(y = anomaly, size = "None", color = "None"), alpha = 0.3) +
  geom_line(aes(y = smooth.1, size = "1-year", color = "1-year")) + 
  geom_line(aes(y = smooth.10, size = "10-year", color = "10-year")) +
  scale_color_manual(values = c("None" = "blue", "1-year" = "green",
                                "10-year" = "blue"), name = "Averaging") +
  scale_size_manual(values = c("None" = 0.1, "1-year" = 0.5,
                               "10-year" = 1.0), name = "Averaging") +
  labs(x = "Year", y = "Temperature Anomaly (Celsius)")
```


The graph shows that temperature didn't show a steady trend until starting
around 1970, so we want to isolate the data starting in 1970 and fit a linear
trend to it.

To select only rows of a tibble that match a condition, we use the function 
`filter` from the `tidyverse` package:

`data_subset = df %>% filter( conditions )`, where `df` is your original tibble
and `conditions` stands for whatever conditions you want to apply.
You can make a simple condition using equalities or inequalities:

* `data_subset = df %>% filter( month == "Jan")` to select all rows where the 
  month is "Jan"
  
* `data_subset = df %>% filter( month != "Aug")` to select all rows where the 
  month is not August.

* `data_subset = df %>% filter( month %in% c("Sep", "Oct", "Nov")` to select all 
  rows where the month is one of "Sep", "Oct", or "Nov".

* `data_subset = df %>% filter(year >= 1945)` to select all rows where the year 
  is greater than or equal to 1945.

* `data_subset = df %>% filter(year >= 1951 & year <= 1980 )` to select all rows
  where the year is between 1951 and 1980, inclusive.

* `data_subset = df %>% filter(year >= 1951 | month == "Mar")` to select all
rows where the year is greater than or equal to 1951 or the month is "Mar". this
will give all rows from January 1951 onward, plus all rows before 1951 where the
month is March.

Below, create a new variable `giss_recent` and assign it a subset of `giss_g`
that has all the data from January 1970 through the present. Fit a linear trend
to the monthly anomaly and report it.

What is the average change in temperature from one year to the next?

```{r giss_trend, include=TRUE}
giss_recent = filter(giss_g, date >= 1970)

recent_trend = lm(anomaly ~ date, data = giss_recent)

tidy(recent_trend)
```

### Did Global Warming Stop after 1998?

It is a common skeptic talking point that global warming stopped in 1998.
In years with strong El Ni&ntilde;os, global temperatures tend to be higher
and in years with strong La Ni&ntilde;as, global temperatures tend to be lower.
We will discuss why later in the semester.

The year 1998 had a particularly strong El Ni&ntilde;o, and the year set a
record for global temperature that was not exceeded for several years. Indeed,
compared to 1998, it might look as though global warming paused for many years.

We will examine whether this apparent pause has scientific validity.

To begin with, we will take the monthly GISS temperature data and convert it to
annual average temperatures, so we can deal with discrete years, rather than 
separate temperatures for each month.

We do this with the `group_by` and `summarize` functions.

We also want to select only recent data, so we arbitrarily say we will look at
temperatures starting in 1979, which gives us 19 years before the 1998 
El Ni&tilde;o.

We don't have a full year of data for 2017, so we want to discard that because
we won't get a full year average from it.

If we go back to the original `giss_g` data tibble, run the following code:

```{r summarize_giss, include=TRUE}
giss_annual = giss_g %>% 
  filter(Year >= 1979 & Year < 2017) %>%
  group_by(Year) %>% 
  summarize(anomaly = mean(anomaly)) %>%
  ungroup() %>%
  mutate(date = Year + 0.5)

head(giss_annual)
```

This code groups the giss data by the year, so that one group will have
January--December 1979, another will have January--December 1980, and
so forth.

Then we replace the groups of 12 rows for each year (each row represents one
month) with a single row that represents the average of those 12 months.

It is important to tell R to `ungroup` the data after we're done working with
the groups.

Finally, we set `date` to `year + 0.5` because the average of a year corresponds
to the middle of the year, not the beginning.

Now, let's introduce a new column `after`, which indicates whether the data is 
after the 1998 El Ni&ntilde;o:

Now plot the data and color the points for 1998 and afterward dark red to
help us compare before and after 1998.

```{r plot_recent_giss, include=TRUE}
ggplot(giss_annual, aes(x = date, y = anomaly)) +
  geom_line(size = 1) +
  # I didn't include it in the original instructions, but the
  # following version of geom_line is nicer than what's above:
  # 
  # geom_line(aes(color = Year >= 1998), size = 1) +
  # 
  geom_point(aes(color = Year >= 1998), size = 2) +
  scale_color_manual(values = c("TRUE" = "red", "FALSE" = "blue"),
                     guide = "none") + # color "before" points blue, 
                                       # "after" points red
                                       # don't use a legend
  labs(x = "Year", y = "Temperature Anomaly (Celsius)")
```

Does it look as though the red points are not rising as fast as the blue points?

Let's just plot the data from 1998 on:

```{r plot_pause, include=TRUE, warning=FALSE}
ggplot(giss_annual, aes(x = date, y = anomaly)) +
  geom_line(size = 1) +
  # I didn't include it in the original instructions, but the
  # following version of geom_line is nicer than what's above:
  # 
  # geom_line(aes(color = Year >= 1998), size = 1) +
  # 
  geom_point(aes(color = Year >= 1998), size = 2) +
  scale_color_manual(values = c("TRUE" = "red", "FALSE" = "blue"),
                     guide = "none") + # color "before" points blue, 
                                       # "after" points red
                                       # don't use a legend
  xlim(1998,2016) +
  labs(x = "Year", y = "Temperature Anomaly (Celsius)")
```

Now how does it look?

Let's use the `filter` function to break the data into two different tibbles: 
`giss_before` will have the data from 1979--1998 and the other, `giss_after` 
will have the data from 1998 onward (note that the year 1998 appears in both
tibbles).

```{r split_giss_data, include=TRUE}
giss_before = filter(giss_annual, Year <= 1998)
giss_after = filter(giss_annual, Year >= 1998)
```

Now use `lm` to fit a linear trend to the temperature data in `giss_before`
(from 1979--1998) and assign it to a variable `giss_trend`.

Next, add a column `timing` to each of the split data sets and set
the value of this column to "Before" for `giss_before` and "After" for
`giss_after`.

```{r add_timing, include=TRUE}
giss_before = mutate(giss_before, timing = "Before")
giss_after = mutate(giss_after, timing = "After")

giss_trend = lm(anomaly ~ date, data = giss_before)
tidy(giss_trend)
```

Now, combine the two tibbles into one tibble:

```{r combine_giss, include=TRUE}
giss_combined <- bind_rows(giss_before, giss_after)
```

Now let's use ggplot to plot `giss_combined`:

* Aesthetic mapping:
    * Use the `date` column for the _x_ variable.
    * Use the `anomaly` column for the _y_ variable.
    * Use the `timing` column to set the color of plot elements
* Plot both lines and points.
    * Set the `size` of the lines to 1
    * Set the `size` of the points to 2
* Use the `scale_color_manual` function to set the color of
  "Before" to "blue" and "After" to "red"
* Use `geom_smooth(data = giss_before, method="lm", color = "blue", fill = "blue", alpha = 0.2, fullrange = TRUE)` to show a linear trend that is fit just to the `giss_before` data.

```{r plot_combined, include = TRUE}
ggplot(giss_combined, aes(x = date, y = anomaly, color = timing)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  geom_smooth(data = giss_before, method = "lm", color = "blue", fill = "blue",
              alpha = 0.2, fullrange = TRUE) +
  scale_color_manual(values = c(Before = "darkblue", After = "red")) +
  labs(x = "Year", y = "Temperature Anomaly (Celsius)")  
```

Try this with the parameter `fullrange` set to `TRUE` and `FALSE` in the 
`geom_smooth` function. What is the difference?

```{r alt_plot_combined_false, include = TRUE}
ggplot(giss_combined, aes(x = date, y = anomaly, color = timing)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  geom_smooth(data = giss_before, method = "lm", color = "blue", fill = "blue",
              alpha = 0.2, fullrange = FALSE) +
  scale_color_manual(values = c(Before = "darkblue", After = "darkred")) +
  labs(x = "Year", y = "Temperature Anomaly (Celsius)")  
```

```{r alt_plot_combined_true, include = TRUE}
ggplot(giss_combined, aes(x = date, y = anomaly, color = timing)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  geom_smooth(data = giss_before, method = "lm", color = "blue", fill = "blue",
              alpha = 0.2, fullrange = TRUE) +
  scale_color_manual(values = c(Before = "darkblue", After = "darkred")) +
  labs(x = "Year", y = "Temperature Anomaly (Celsius)")  
```

**Answer:** 
Both plots show the full data set, and a linear trend that is fit just 
to the "before" data. The trend line shows both the best fit for a trend
(that's the solid line) and the range of uncertainty in the fit (that's the 
light blue shaded area around the line).

But, when `fullrange = FALSE`, the line is only drawn for the data to which the
trend was fit, whereas when `fullrange = TRUE`, the trend line is drawn for the
full range of the graph, even though the trend was only fit to the data in part
of the graph.

If the temperature trend changed after 1998 (e.g., if the warming paused, or if
it reversed and started cooling) then we would expect the temperature
measurements after 1998 to fall predominantly below the extrapolated trend line,
and our confidence that the trend had changed would depend on the number of
points that fall below the shaded uncertainty range.

How many of the red points fall below the trend line?

**Answer:** 6 points: 1999, 2000, 2008, 2011, 2012, and 2013.

How many of the red points fall above the trend line?

**Answer:** Not counting 1998, 12 points: 2001, 2002, 2003, 2004 (barely), 
2005, 2006, 2007, 2009, 2010, 2014, 2015, and 2016.

What do you conclude about whether global warming paused or stopped after 1998?

**Answer:** Most of the years after 1998 were warmer than we would have
predicted if temperatures had continued to follow the warming trends of
1979--1998, so this is not evidence of any slow-down or pause in the warming.

