---
title: 'Lab #3 Answers'
lab_number: 3
lab_date: '2022-02-07'
sol_date: '2022-02-20'
pubdate: '2022-02-20'
date: '2022-02-16'
pdf_url: /files/lab_solutions/lab_03/lab_03_answers.pdf
slug: lab_03_answers
output:
  blogdown::html_page:
    md_extensions: +tex_math_single_backslash+compact_definition_lists
    toc: yes
---
```{r setup, include=FALSE}
knitr::knit_hooks$set(inline = function(x) { knitr:::format_sci(x, 'md')})
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)

if (knitr::is_latex_output()) {
  message("Setting PDF plot hook.")
  new_plot_hook <- function(x, options)
    paste("\n", knitr::hook_plot_tex(x, options), "\n")
  
  knitr::knit_hooks$set(plot = new_plot_hook)
} else {
  message("Not using PDF output.")
}


options(dplyr.summarise.inform = FALSE)

# This section loads necessary R libraries and sources scripts that define 
# useful functions format_md.
# 
data_dir = "_data"
script_dir = "_scripts"

library(pacman)
p_load(tidyverse, knitr)

theme_set(theme_bw(base_size = 15))

source(file.path(script_dir, 'utils.R'), chdir = T)
source(file.path(script_dir, 'format_md.R'), chdir = T)
source(file.path(script_dir, 'modtran.R'), chdir = T)

```
# Chapter 4 Exercises

Fill in R code for the exercises 
(I have put the comment `# TODO` in all of the code chunks where you need to 
do this)
and then fill in the answers where I have marked **Answer:**. 
Be sure to write explanations of your answer and don't just put numbers with 
no text.

## Exercise 4.1: Methane

Methane has a current concentration of 1.7 ppm in the atmosphere and
is doubling at a faster rate than CO~2~.

a) **Would an additional 10 ppm of methane in the atmosphere have a larger or 
   smaller impact on the outgoing IR flux than an additional 10 ppm of CO~2~ 
   at current concentrations?**

   **Hint:**
   See the suggestion in the `lab-03-instructions` document.

```{r ex_4_1_a, include=TRUE, message=FALSE}
base_co2 = 400 # parts per million
base_ch4 = 1.7 # parts per million
modtran_baseline = run_modtran(file.path(data_dir, 
                                         "ex_4_1_baseline.txt"),
                               co2_ppm = base_co2, 
                               ch4_ppm = base_ch4)
modtran_plus_10_co2 = run_modtran(file.path(data_dir, 
                                            "ex_4_1_plus_10_co2.txt"),
                                  co2_ppm = base_co2 + 10, 
                                  ch4_ppm = base_ch4)
modtran_plus_10_ch4 = run_modtran(file.path(data_dir, 
                                            "ex_4_1_plus_10_ch4.txt"),
                                  co2_ppm = base_co2, 
                                  ch4_ppm = base_ch4 + 10)
i_out_baseline = modtran_baseline$i_out
i_out_co2 = modtran_plus_10_co2$i_out
i_out_ch4 = modtran_plus_10_ch4$i_out
delta_i_co2 = i_out_co2 - i_out_baseline
delta_i_ch4 = i_out_ch4 - i_out_baseline
```

**Answer:** 
I ran MODTRAN three times. One run was a baseline, which used the 
current concentrations of CO~2~ and CH~4~. Then I ran MODTRAN with the CO~2~
concentration increased by 10 ppm and I ran it a third time with the baseline 
value for CO~2~, but with CH~4~ increased by 10 ppm.

For the baseline run, the intensity of outgoing longwave light
was `r format_md(i_out_baseline, digits = 3)`. Watts per square meter.
Increasing CO~2~ by 10 ppm decreased the outgoing longwave light by 
`r format_md(-delta_i_co2, digits = 1)` W/m^2^, and increasing CH~4~ by 
10 ppm decreased the outgoing longwave light by 
`r format_md(-delta_i_ch4, digits = 2)` W/m^2^, which is around 
`r format_md(delta_i_ch4 / delta_i_co2, digits = 1)` times as much as for CO~2~.

The difference is because absorption for CO~2~ is strongly saturated, but the 
absorption for CH~4~ is not saturated. Another way to think about this is that
a 10 ppm increase in CO~2~ increases the atmospheric concentration by 
`r format_md(100 * 10 / base_co2, digits = 1)`% and a 10 ppm increase in 
CH~4~ increases the atmospheric concentration by 
`r format_md(100 * 10 / base_ch4, digits = 1)`%.

b) **Where in the spectrum does methane absorb? What concentration does it take 
   to begin to saturate the absorption in this band? Explain what you are 
   looking at to judge when the gas is saturated.**

   **Hints:**  
   See the hints in the `lab-03-instructions` document.

```{r ex_4_1_b, include=TRUE, message=FALSE}
ch4_values = c(0, 2^seq(0,7))
sat_data = tibble()
for (ch4 in ch4_values) {
   filename = file.path(data_dir, str_c("ex_4_1_ch4_", ch4, ".txt"))
   if (file.exists(filename)) {
      mod_data = read_modtran(filename)
   } else {
      mod_data = run_modtran(filename, co2_ppm = 0, ch4_ppm = ch4, 
                             trop_o3_ppb = 0, strat_o3_scale = 0, 
                             h2o_scale = 0, freon_scale = 0,
                             atmosphere = "standard", altitude_km = 70)
   }
   p = plot_modtran(mod_data, descr = str_c(ch4, " ppm methane"))
   plot(p)
}
```

**Answer:** 
The plots above show the MODTRAN spectrum with all gases set to zero except
methane. Methane absorbs most strongly around 1300 wavenumbers.

As we increase the methane concentration, the big spike around 1300 wavenumbers
gets bigger until it bottoms out on the black line. This happens somwhere around
8, 16, or 32 ppm, so any of those anwers would be correct. But the spectrum is 
complicated and so is its saturation, so other answers are plausible if they
are supported by sound reasoning.
  
c) **Would a doubling of methane have as great an impact on the heat balance as
   a doubling of CO~2~?**

   **Hint:**
   See the suggestion in the `lab-03-instructions` document.

```{r ex_4_1_c, include=TRUE, message=FALSE}
base_co2_ppm = 400
base_methane_ppm = 1.7
baseline =   read_modtran(file.path(data_dir, "ex_4_1_baseline.txt"))
double_co2 = run_modtran(file.path(data_dir, "ex_4_1_2x_co2.txt"),
                         co2_ppm = 2 * base_co2_ppm, 
                         ch4_ppm = base_methane_ppm)
double_ch4 = run_modtran(file.path(data_dir, "ex_4_1_2x_ch4.txt"),
                         co2_ppm = base_co2_ppm, 
                         ch4_ppm = 2 * base_methane_ppm)
i_out_baseline = baseline$i_out
i_out_co2 = double_co2$i_out
i_out_ch4 = double_ch4$i_out

delta_i_out_co2 = i_out_co2 - i_out_baseline
delta_i_out_ch4 = i_out_ch4 - i_out_baseline
```

**Answer:** 
The baseline value for I~out~ is 
`r format_md(i_out_baseline, digits = 3)` W/m^2^. 
If we double CO~2~, it drops to
`r format_md(i_out_co2, digits = 3)` W/m^2^, a decrease of
`r format_md(delta_i_out_co2, digits = 1)` W/m^2^, 
and if we double CH~4~, it drops to 
`r format_md(i_out_ch4, digits = 3)` W/m^2^, a decrease of
`r format_md(delta_i_out_ch4, digits = 1)` W/m^2^.
Doubling CO~2~ has the larger effect because there is a lot more CO~2~ in the 
atmosphere and that is more important than how saturated its absorption is.

You can see this if you look at the plots below. Notice that the effect of 
doubling CO~2~ isn't to make the big CO~2~ absorption feature get deeper, but 
to make it wider. You can see this if you compare the baseline to the doubled
CO~2~ spectrum where the purple spectrum crosses the yellow blackbody curve
near 600 wavenumbers and around 750 wavenumbers.  Compare this to the very 
small change in the methane spike near 1300 wavenumbers (you have to look 
very carefully at the doubled methane spectrum to notice this).

```{r ex_4_1_c_baseline_spectrum}
plot_modtran(baseline, descr = "Baseline spectrum")
```
```{r ex_4_1_c_double_co2_spectrum}
plot_modtran(double_co2, descr = "Doubled CO2")
```
```{r ex_4_1_c_double_ch4_spectrum}
plot_modtran(double_ch4, descr = "Doubled CH4")
```



d) **What is the "equivalent CO~2~" of doubling atmospheric methane? That is to 
   say, how many ppm of CO~2~ would lead to the same change in outgoing IR 
   radiation energy flux as doubling methane? What is the ratio of ppm CO~2~ 
   change to ppm methane change?**

```{r ex_4_1_d, include=TRUE, message=FALSE}
matching_methane = 13
modtran_match_ch4 = run_modtran(file.path(data_dir, "ex_4_1_ch4_match.txt"),
                                co2_ppm = 400, ch4_ppm = matching_methane)
```

**Answer:** 
When we double CO~2~, I~out~ is 
`r `format_md(i_out_co2, digits = 4)`W/m^2^.
We need to adjust CH~4~ to produce the same I~out~ with the defailt value of
400 ppm CO~2~. After some trial and error, this turns out to be about 
`r format(matching_methane, digits = 1)` ppm, 
which has I~out~ = `r format_md(modtran_match_ch4$i_out, digits = 4)`

## Exercise 4.2: CO~2~ (Graduate students only)

a) **Is the direct effect of increasing CO~2~ on the energy output at the top of
   the atmosphere larger in high latitudes or in the tropics?**

   **Hint:**
   See the hint in the `lab-03-instructions` document.


```{r ex_4_2_a, include=TRUE, message=FALSE}
modtran_lat_df = tibble()
for (atmos in c("tropical", "midlatitude summer", "subarctic summer"
                #, "midlatitude winter", "subarctic winter"
)) {
  for (co2 in c(400, 800)) {
    filename = file.path(data_dir, 
                         str_c("ex_4_2_", atmos, "_co2_", co2, ".txt"))
    if (file.exists(filename)) {
      mod_data = read_modtran(filename)
    } else {
      mod_data = run_modtran(filename, atmosphere = atmos, co2_ppm = co2)
    }
    modtran_lat_df = bind_rows(modtran_lat_df, 
                               tibble(co2 = co2, atmos = atmos, 
                                      i_out = mod_data$i_out))
  }
}

modtran_lat_df = modtran_lat_df %>% 
  mutate(co2 = str_c("co2_", co2)) %>%
  pivot_wider(names_from = "co2", values_from = "i_out") %>%
  mutate(delta_i = co2_400 - co2_800) %>%
  arrange(desc(delta_i))
kable(modtran_lat_df, digits = 1)
```

**Answer:** 
When atmospheric CO~2~ doubles, the greatest change in I~out~ occurs
in the tropics, followed by the midlatitudes, and the smallest change occurs at
high latitudes in the subarctic.

b) **Set pCO~2~ to an absurdly high value of 10,000 ppm. You will see a spike
   in the CO~2~ absorption band. What temperature is this light coming from? 
   Where in the atmosphere do you think this comes from?**

   **Now turn on clouds and run the model again. Explain what you see.
   Why are night-time temperatures warmer when there are clouds?**

   **Hint:**
   See the hint in the `lab-03-instructions` document and
   for the second part of this exercise, try using "altostratus" clouds.

The figure below shows the spectrum with a high concentration of CO~2~.

```{r ex_4_2_b_altitude, include=TRUE, message=FALSE}
high_co2 = 1E4
for (z in c(70, 60, 50, 40, 30, 20, 10)) {
  filename = file.path(data_dir, str_c("ex_4_2_alt_", z, ".txt"))
  if (file.exists(filename)) {
    modtran_high_co2 = read_modtran(filename)
  } else {
    modtran_high_co2 = run_modtran(filename, co2_ppm = high_co2, 
                                   altitude_km = z)
  }
  p = plot_modtran(modtran_high_co2)
  plot(p)
}
```

```{r ex_4_2_b_clear, include=TRUE, message=FALSE}
high_co2 = 1E4
modtran_high_co2 = run_modtran(file.path(data_dir, 
                                         "ex_4_2_hi_co2_clear.txt"), 
                               co2_ppm = high_co2)
plot_modtran(modtran_high_co2)
```


```{r ex_4_2_b_clouds, include=TRUE, message=FALSE}
high_co2 = 1E4
modtran_high_co2 = run_modtran(file.path(data_dir, 
                                         "ex_4_2_hi_co2_cloudy.txt"),
                               co2_ppm = high_co2, clouds = "altostratus")
plot_modtran(modtran_high_co2)
```


**Answer:** 
The spike in the CO~2~ absorption feature gets smaller as the sensor
altitude drops below about 40 km, and disappears entirely as the sensor drops
below 30 km. This means that the spike must be coming from the region of the 
atmosphere between 30 and 40 km.

When you add altostratus clouds to the atmosphere there isn't much change in
the emission from the big absorption spikes but the emission from the window
region of the spectrum drops considerably. The clouds block longwave emissions
in the window region. This traps heat near the surface and is one reason why
cloudy nights tend to be warmer than clear nights.

## Exercise 4.3: Water vapor

Our theory of climate presumes that an increase in the temperature at ground
level will lead to an increase in the outgoing IR energy flux at the top of the
atmosphere.


a) **How much extra outgoing IR would you get by raising the temperature of the 
   ground by 5&deg;C? What effect does the ground temperature have on the 
   shape of the outgoing IR spectrum and why?**
   
   **Hint:**
   See the hint in the `lab-03-instructions` document.

```{r ex_4_3_a, include=TRUE, message=FALSE}
modtran_baseline = run_modtran(file.path(data_dir, "ex_4_3_baseline.txt"))
modtran_plus_5 = run_modtran(file.path(data_dir, "ex_4_3_t_plus_5"), 
                             delta_t = 5)

p_baseline = plot_modtran(modtran_baseline, descr = "Baseline spectrum")
p_plus_5 = plot_modtran(modtran_plus_5, 
                        descr = "Ground temperature raised 5K")
plot(p_baseline)
plot(p_plus_5)
```

**Answer:** 
Raising the ground temperature raises the entire spectrum.

b) **More water can evaporate into warm air than into cool air. Change the
   model settings to hold the water vapor at constant relative humidity 
   rather than constant vapor pressure (the default), calculate the change
   in outgoing IR energy flux for a 5&deg;C temperature increase.
   Is it higher or lower? Does water vapor make the Earth more sensitive to
   CO~2~ increases or less sensitive?**
   
   **Note:** By default, the MODTRAM model holds water vapor pressure constant,
   but you can set it to hold relative humidity constant instead with the
   option `h2o_fixed = "relative humidity"`, like this:
   `run_modtran(file_name, delta_t = 5, h2o_fixed = "relative humidity")`.

```{r ex_4_3_b, include=TRUE, message=FALSE}
modtran_vp = run_modtran(file.path(data_dir, "ex_4_3_vap_press.txt"),
                         delta_t = 5, h2o_fixed = "vapor pressure")
modtran_rh = run_modtran(file.path(data_dir, "ex_4_3_rel_humid.txt"),
                         delta_t = 5, h2o_fixed = "relative humidity")

i_out_baseline = modtran_baseline$i_out
i_out_vp = modtran_vp$i_out
i_out_rh = modtran_rh$i_out

p_vp = plot_modtran(modtran_vp, descr = "Constant vapor pressure", 
                    i_out_ref = i_out_baseline)
p_rh = plot_modtran(modtran_rh, descr = "Constant relative humidity", 
                    i_out_ref = i_out_baseline)
plot(p_vp)
plot(p_rh)
```

**Answer:** 
Raising the surface temperature has a bigger effect on I~out~ when
water vapor pressure is fixed than when relative humidity is fixed. This means
that compensating for a change in CO~2~ would require a bigger change in 
temperature with fixed relative humidity, so the climate is more sensitive to
changes in CO~2~ when relative humidity is fixed.

c) **Now see this effect in another way.**

   * **Starting from the default base case, record the total outgoing 
     IR flux.**

   * **Now double CO~2~. The temperature in the model stays the
     same (that's how the model is written), but the outgoing IR flux
     goes down.**

   * **Using constant water vapor pressure, adjust the temperature offset
     until you get the original IR flux back again. Record the change in
     temperature.**
    
   * **Now repeat the exercise, but holding the relative humidity fixed
     instead of the water vapor pressure.**
    
   * **The ratio of the warming when you hold relative humidity fixed
     to the warming when you hold water vapor pressure fixed is the 
     feedback factor for water vapor. What is it?**

```{r ex_4_3_c, include=TRUE, message=FALSE}
modtran_baseline = read_modtran(file.path(data_dir, "ex_4_3_baseline.txt"))
i_baseline = modtran_baseline$i_out

modtran_2xco2 = run_modtran(file.path(data_dir, "ex_4_3_2x_co2_vp.txt"),
                                      co2_ppm = 800)
i_2x_vp = modtran_2xco2$i_out

dt_vp = 0.76
modtran_vp_dt = run_modtran(file.path(data_dir, "ex_4_3_vp_dt.txt"),
                            co2_ppm = 800, delta_t = dt_vp)
i_vp_dt = modtran_vp_dt$i_out

modtran_2x_rh = run_modtran(file.path(data_dir, "ex_4_3_2x_co2_rh_.txt"),
                            co2_ppm = 800, h2o_fixed = "relative humidity")
i_2x_rh = modtran_2x_rh$i_out

dt_rh = 1.21
modtran_rh_dt = run_modtran(file.path(data_dir, "ex_4_3_rh_dt.txt"),
                            co2_ppm = 800, delta_t = dt_rh, 
                            h2o_fixed = "relative humidity")
i_rh_dt = modtran_rh_dt$i_out

feedback_factor = dt_rh / dt_vp
```

**Answer:** 
In the baseline case, 
I~out~ = `r format_md(i_baseline, digits = 4)` W/m^2^. 
When we double CO~2~ with constant water vapor pressure, I~out~ drops to 
`r format_md(i_2x_vp, digits = 4)` W/m^2^ 
and we have to raise the ground temperature by 
`r format_md(dt_vp, digits = 2)` K 
to bring I~out~ back to 
I~out~ = `r format_md(i_vp_dt, digits = 4)` W/m^2^. 

When we double CO~2~ with constant water relative humidity, I~out~ drops to 
`r format_md(i_2x_rh, digits = 4)` W/m^2^ 
and we have to raise the ground temperature by 
`r format_md(dt_rh, digits = 2)` K 
to bring I~out~ back to 
I~out~ = `r format_md(i_rh_dt, digits = 4)` W/m^2^. 

The feedback factor is the ratio of the temperature change with relative 
humidity fixes to the temperature change with vapor pressure fixed:
_f_ = `r format_md(feedback_factor, digits = 2)`.

Notice that there is no difference between holding vapor pressure constant and 
holding relative humidity constant until the temperature changes.

# Chapter 5 Exercise

## Exercise 5.2: Skin Height

a) **Run the MODTRAN model using the "tropical" atmosphere, without clouds, 
   and with the present-day CO~2~ concentration (400 ppm). 
   Use the ground temperature reported 
   by the model to calculate
   $\varepsilon \sigma T_{\text{ground}}^4$,
   the heat flux emitted by the ground.
   Assume $\varepsilon = 1$, and I have already provided the value of the 
   Stefan-Boltzmann constant $\sigma$, as the R variable `sigma_sb`,
   which equals `r format_md(sigma_sb, digits = 3, format = "scientific")`.
   (I defined it in the script "utils.R", which I loaded in the "setup" chunk
   in the RMarkdown document).**

   **Next, look at the outgoing heat flux at the top of the atmosphere (70 km) 
   (I~out~) reported by the MODTRAN model. Is it greater or less than the heat 
   flux that you calculated was emitted by the ground?**

```{r ex_5_2_a, include=TRUE, message=FALSE}
baseline = run_modtran(file.path(data_dir, "ex_5_2_baseline.txt"),
                       co2_ppm = 400, atmosphere = "tropical", 
                       clouds = "none")
epsilon = 1.0
t_ground = baseline$t_ground
i_ground = epsilon * sigma_sb * t_ground^4
i_atmos_up = baseline$i_out
```

**Answer:** 
T~ground~ = `r format_md(t_ground, digits = 3)` K, so 
$I_{\text{out}} = \varepsilon \sigma T_{\text{ground}}^4 = 
`r format_md(i_ground, digits = 3)`$ W/m^2^. The MODTRAN model reports 
I~out~ = `r format_md(i_atmos_up, digits = 3)` W/m^2^. 
I~ground~ is roughly `r format_md(i_ground / i_atmos_up, digits = 1)` times
greater than I~out~.

b) **Use the outgoing heat flux at the top of the atmosphere (I~out~) to 
   calcuate the skin temperature (use the equation 
   $I_{\text{out}} = \varepsilon \sigma T_{\text{skin}}^4)$).
   What is the skin temperature, and how does it compare to the ground 
   temperature and the temperature at the tropopause, as reported by the 
   MODTRAN model (`t_tropo`)?**
   
   **Assuming an environmental lapse rate of 6K/km, and using the 
   skin temperature that you calculated above, and the ground temperature
   from the model, what altitude would you expect the skin height to be?**

```{r ex_5_2_b, include=TRUE, message=FALSE}
lapse_rate = 6.0 # Kelvin/km
t_tropo = baseline$t_tropo
t_skin = (i_atmos_up / (epsilon * sigma_sb))^0.25
h_skin = (t_ground - t_skin) / lapse_rate
```

**Answer:** The skin temperature is given by 

$$T_{\text{skin}} = \sqrt[4]{\frac{I_{\text{out}}}{\varepsilon \sigma}}$$
T~skin~ = `r format_md(t_skin, digits = 3)` K, which is
`r format_md(t_ground - t_skin, digits = 2)` K less than the ground temperature
and greater than the tropopause temperature 
T~tropo~ = `r format_md(t_tropo, digits = 3)`.
The skin height is 
$$h_{\text{skin}} = \frac{T_{\text{ground}} - T_{\text{skin}}}{\text{lapse rate}}$$
so with a lapse rate of `r lapse_rate` K/km, 
h~skin~ = `r format_md(h_skin, digits = 1)` km.

c) **Double the CO~2~ concentration and run MODTRAN again. Do not adjust the
   ground temperature. Repeat the calculations from (b) of the skin
   temperature and the estimated skin height.**

   **What is the new skin temperature? What is the new skin height?**

```{r ex_5_2_c, include=TRUE, message=FALSE}
modtran_2x_co2 = run_modtran(file.path(data_dir, "ex_5_2_2x_co2_clear.txt"),
                             co2_ppm = 800, atmosphere = "tropical",
                             clouds = "none")
i_out_2x = modtran_2x_co2$i_out
t_skin_2x = (i_out_2x / (epsilon * sigma_sb))^0.25
h_skin_2x = (t_ground - t_skin_2x) / lapse_rate
```

**Answer:** 
The new I~out~ is `r format_md(i_out_2x, digits = 3)` W/m^2^,
so the new T~skin~ is `r format_md(t_skin_2x, digits = 3)` K, which implies
that the new skin height is `r format_md(h_skin_2x, digits = 1)` km,
`r format_md(h_skin_2x - h_skin, digits = 1)` km higher than for today's 
CO~2~ concentration.

d) **Put the CO~2~ back to today's value, but add cirrus clouds, using the
   "standard cirrus" value for the clouds. Repeat the calculations from (b) of
   the skin temperature and the skin height.**
   
   **What is the new skin temperature? What is the new skin height?
   Did the clouds or the doubled CO~2~ have a greater effect on the
   skin height?**

```{r ex_5_2_d, include=TRUE, message=FALSE}
modtran_cirrus = run_modtran(file.path(data_dir, "ex_5_2_cirrus.txt"),
                             co2_ppm = 400, atmosphere = "tropical", 
                             clouds = "standard cirrus")
i_out_cirrus = modtran_cirrus$i_out
t_skin_cirrus = (i_out_cirrus / (epsilon * sigma_sb))^0.25
h_skin_cirrus = (t_ground - t_skin_cirrus) / lapse_rate
```

**Answer:** 
The new skin height is `r format_md(h_skin_cirrus, digits = 1)` km, which is 
`r format_md(h_skin_cirrus - h_skin_2x, digits = 1)` km higher than for the 
doubled CO~2~. To put this in context, doubling CO~2~ raises the skin hight by
`r format_md(h_skin_2x - h_skin, digits = 1)` km and adding cirrus clouds raises
the skin height by `r format_md(h_skin_cirrus - h_skin, digits = 1)`, 
so the cirrus clouds have a much bigger effect on the climate.
