---
title: 'Lab #4 Answers'
lab_number: 4
lab_date: '2021-02-15'
sol_date: '2021-02-11'
pubdate: '2021-02-11'
date: '2021-03-01'
pdf_url: /files/lab_solutions/lab_04/lab_04_answers.pdf
slug: lab_04_answers
output:
  blogdown::html_page:
    md_extensions: +tex_math_single_backslash+compact_definition_lists
    toc: yes
---
```{r setup, include=FALSE}
knitr::knit_hooks$set(inline = function(x) { knitr:::format_sci(x, 'md')})
knitr::opts_chunk$set(echo = TRUE, include = TRUE, cache = FALSE)

if (knitr::is_latex_output()) {
  message("Setting PDF plot hook.")
  new_plot_hook <- function(x, options)
    paste("\n", knitr::hook_plot_tex(x, options), "\n")

  knitr::knit_hooks$set(plot = new_plot_hook)
} else {
  message("Not using PDF output.")
}


options(dplyr.summarise.inform = FALSE)

# This section loads necessary R libraries and sources scripts that define
# useful functions format_md.
#
data_dir = "_data"
script_dir = "_scripts"

library(pacman)
p_load(tidyverse, jsonlite, httr, magrittr, scales, xml2, knitr)

theme_set(theme_bw(base_size = 10))

source(file.path(script_dir, "utils.R"), chdir = T)
source(file.path(script_dir, "format_md.R"), chdir = T)
source(file.path(script_dir, "modtran.R"), chdir = T)
source(file.path(script_dir, "rrtm.R"), chdir = T)
```
# General Instructions

In the past three weeks, we focused on mastering many of the basics of using
R and RMarkdown. For this week's lab, when you write up the answers, I would
like you to think about integrating your R code chunks with your text.

For instance, you can describe what you're going to do to answer the question,
and then for each step, after you describe what you're going to do in that
step, you can include an R code chunk to do what you just described, and then
the subsequent text can either discuss the results of what you just did
or describe what the next step of the analysis will do.

This way, your answer can have several small chunks of R code that build on
each other and follow the flow of your text.

For this lab, you will use the RRTM model, which includes both radiation and
convection.

## Exercise 1: Lapse Rate

Run the RRTM model in its default configuration and then vary the lapse rate
from 0 to 10 K/km. For each value of the lapse rate, adjust the surface
temperature until the earth loses as much heat as it gains (i.e., the value of
_Q_ in the `run_rrtm` model output is zero.)

It will probably be easier to do this with the interactive version of the RRTM
model at <http://climatemodels.uchicago.edu/rrtm/> than with the R interface
`run_rrtm`.

a) Make a tibble containing the values of the lapse rate and the corresponding
   equilibrium surface temperature, and make a plot with lapse rate on the
   horizontal axis and surface temperature on the vertical axis.


b) Describe how the equilibrium surface temperature varies as the lapse rate
   varies.

### Exercise 1 Report:

I varied the lapse rate from 0 to 10 Kelvin/km in steps of 1 K/km.
At each value of the lapse rate, I manually adjusted the surface temperature in
the interactive RRTM model until the heat budget was balanced (the model 
reported that "If the Earth has these properties ... then it loses as much
energy as it gains."). 

```{r ex-1-a, include=TRUE, echo=TRUE, warning=FALSE, message=FALSE}
lapse_vs_t = tibble( lapse = seq(0, 10),
                     t_surf = c(256.05, 264.66, 271.65, 276.95, 280.75, 283.15,
                                284.40, 285.00, 285.25, 285.30, 285.30))
kable(lapse_vs_t, col.names = c("Lapse rate", "Surface temperature"))
```

The figure below shows the results. As the lapse rate increases, the change in 
surface temperature becomes smaller and smaller. At the last step,
from 9 to 10 K/km, the surface temperature did not change at all.

```{r ex-1-a-plot, include=TRUE, echo=TRUE, warning=FALSE, message=FALSE, fig.cap="Equilibrium surface temperature versus lapse rate."}
ggplot(lapse_vs_t, aes(x = lapse, y = t_surf)) +
   geom_line(size = 1, color = "darkblue") +
   geom_point(size = 3, color = "darkblue") +
   labs(x = "Lapse rate (K/km)", y = "Ground temperature (K)")
```

#### **Explanation:**

Students don't need to have this detail in their reports, but the reason this 
is happening is that the closer the basic environmental lapse rate gets to the
dry adiabatic lapse rate, the less stable the atmosphere is and the easier it
is for solar heating of the surface to set off convection that redistributes
heat (basically bringing heat from the surface to the upper troposphere).

When the environmental lapse rate (ELR) is less than the moist adiabatic lapse 
rate, the surface temperature is close to 
$T_{\text{surface}} = T_{\text{skin}} + h_{\text{skin}} \text{ELR}$,
but when the ELR becomes greater than the moist adiabatic lapse rate, there
is so much convection that it disrupts the simple picture I presented in class.

The picture I presented in class works well for an atmosphere like ours, which 
is marginally stable (i.e., the environmental lapse rate is roughly equal to the
average moist adiabatic lapse rate).
When the atmosphere is undergoing a large amount of constant convection, then 
heat flow behaves differently.

All of that convection transports lots of heat from the surface to the upper
troposphere and brings lots of cold air down from the upper troposphere to the
surface, which makes it hard for the surface to get warmer. This is why the
ground temperature flattens out and stops changing very much after the lapse
rate reaches about 6 K/km.

# Exercises on Albedo and Clouds

For the following exercises, start off with the RRTM model in its default 
configuration. 
Record the ground temperature.
For each part of this exercise you will do the following:

1. You will adjust the albedo or the clouds.
2. You will compare the visible and longwave radiation going down through the 
   atmosphere to the surface and also the visible and longwave radiation 
   going up from the surface, through the atmosphere, to space.
   
   The results of an RRTM model run have a tibble called `fluxes` with columns
   for `altitude`, `sw_down`, `sw_up`, `lw_down`, `lw_up`, `total_down`, and
   `total_up`, where `sw` means shortwave, `lw` means longwave, and `total`
   is the sum of shortwave plus longwave.
   
   The first row of this tibble is at ground-level and the last row
   is at the top of the atmosphere.
   
   ``` r 
   default_data = run_rrtm()
   fluxes = default_data$fluxes
   surface_fluxes = head(fluxes, 1) # get the first row
   space_fluxes = tail(fluxes, 1)   # get the last row
   ```
   
3. You will adjust the ground temperature until the heat coming in balances
   the heat going out (the model will say, 
   "If the Earth has these properties ... then it loses as much energy as it
   gains."

## Exercise 2: The urban heat island

First, run the RRTM model in its default configuration and note the surface
temperature and the albedo.

a) Change the surface type from "Earth's average" to "Asphalt" (don't change
   the surface temperature until the instructions tell you to) and describe
   the changes in the local climate:
   * What is the albedo?
   * Report the changes in shortwave and longwave light absorbed by 
     the surface and going out to space.
   * How much does the total balance of heat change (i.e., how many W/m^2^ does
     the Earth lose or gain)?
   * Now, adjust the ground temperature until the Earth loses as much energy as 
     it gains. 
   * What is the new surface temperature? 
     How does it compare to the surface temperature in the default 
     configuration?

b) Change the surface albedo to "Concrete". Answer the same questions as in part
   (a).
   
c) In cities, streets and parking lots are usually paved with asphalt.
   Roofs of houses and other buildings are often covered with asphalt shingles 
   or black rubber-like compounds. 

   The results you got in this exercise represent covering _the entire planet_
   with asphalt or concrete, so they are far more extreme than you would get
   from only covering _part of a city_ with one material or the other, but 
   the general principle holds and in a city you would have much smaller 
   changes, but they would be in the same direction as you found here.

   How would the choice of using asphalt for roads, parking lots, and roofs in
   a large city affect the local climate in the city?
   Would using low-albedo materials, such as concrete for streets and parking 
   lots and light-colored polymers for the roofs of buildings have a benefit 
   for the people living in the city?

### Exercise 2 Report:

#### **Preliminaries**

Because I will be doing a lot of comparisons here, I want to automate some of
the text using a `compare` function:
```{r compare-function}
compare = function(x, y) {
   ifelse(x > y, "greater than", 
          ifelse(x < y, "less than", "the same as"))
}

# This function compares two values and prints the absolute value of 
# the difference followed by optional units (e.g., W/m^2^), and then the
# comparison string, so something like "3.2 W/m^2^ less than".
#
# If the two values are the same, we don't want to print 
# "0 W/m^2^ the same as", so we suppress printing the number and the units 
# in that case.
compare_val = function(x, y, digits, units = NULL) {
   delta = format_md(abs(x - y), digits = digits)
   comp_str = compare(x, y)
   ifelse(x == y, comp_str, str_c(delta, units, comp_str, sep = " "))
}
```

First, I recorded the albedo, surface temperature, and heat fluxes in and out
of the earth under the default conditions of the earth's average albedo.

```{r RRTM_default}
default = run_rrtm()
albedo_default = default$albedo
T_surface_default = default$T_surface

I_in_default = default$i_in
I_out_default = default$i_out

flux_surf_default = default$fluxes %>% filter(altitude == 0)
flux_space_default = tail(default$fluxes, 1)

I_surf_long_default = flux_surf_default$lw_down
I_surf_short_default = flux_surf_default$sw_down
I_space_long_default = flux_space_default$lw_up
I_space_short_default = flux_space_default$sw_up
```

The albedo was 
`r format_md(albedo_default, digits = 1)`,
I~in~ = `r format_md(I_in_default, digits = 3)` W/m^2^, and
I~out~ = `r format_md(I_out_default, digits = 3)` W/m^2^, so the heat 
budget is balanced.

Breaking down the heat flow into shortwave and longwave components, the 
heat absorbed by the ground consisted of
`r format_md(I_surf_long_default, digits = 3)` W/m^2^ of longwave radiation and
`r format_md(I_surf_short_default, digits = 3)` W/m^2^ of shortwwave radiation.
The heat radiated out to space consisted of
`r format_md(I_space_long_default, digits = 3)` W/m^2^ of longwave radiation and
`r format_md(I_space_short_default, digits = 3)` W/m^2^ of shortwwave radiation.


#### 2(a) **Asphalt Surface**

Next, I changed the surface type to "Asphalt" and recorded the changes in 
albedo, I~in~, I~out~, and heat balance.

```{r ex-2-a, include=TRUE, warning=FALSE, message=FALSE}
asphalt = run_rrtm(surface_type = "Asphalt")
albedo_asphalt = asphalt$albedo

I_in_asphalt = asphalt$i_in
I_out_asphalt = asphalt$i_out

# This is the net heat flow into the earth.
# It should be zero when heat in = heat out
Q_asphalt = asphalt$Q

flux_surf_asphalt = asphalt$fluxes %>% filter(altitude == 0)
flux_space_asphalt = tail(asphalt$fluxes, 1)

I_surf_long_asphalt = flux_surf_asphalt$lw_down
I_surf_short_asphalt = flux_surf_asphalt$sw_down
I_space_long_asphalt = flux_space_asphalt$lw_up
I_space_short_asphalt = flux_space_asphalt$sw_up

Delta_I_in_asphalt = I_in_asphalt - I_in_default
Delta_I_out_asphalt = I_out_asphalt - I_out_default

Delta_I_surf_long_asphalt = I_surf_long_asphalt - I_surf_long_default
Delta_I_surf_short_asphalt = I_surf_short_asphalt - I_surf_short_default
Delta_I_space_long_asphalt = I_space_long_asphalt - I_space_long_default
Delta_I_space_short_asphalt = I_space_short_asphalt - I_space_short_default
```

The albedo for asphalt is `r format_md(albedo_asphalt, digits = 0)`, 
which is a good deal `r compare(albedo_asphalt, albedo_default)` the default 
albedo. 
This caused I~in~ to increase from 
`r format_md(I_in_default, digits = 3)` W/m^2^ to
`r format_md(I_in_asphalt, digits = 3)` W/m^2^.
I~out~ did not change and remained at 
`r format_md(I_out_asphalt, digits = 3)` W/m^2^. 

Breaking down the heat flow into shortwave and longwave components, the 
heat absorbed by the ground consisted of
`r format_md(I_surf_long_asphalt, digits = 3)` W/m^2^ of longwave radiation,
which was 
`r compare_val(I_surf_long_asphalt, I_surf_long_default, digits = 2, units = "W/m^2^")`
in the default case, and
`r format_md(I_surf_short_asphalt, digits = 3)` W/m^2^ of shortwwave radiation,
which was 
`r compare_val(I_surf_short_asphalt, I_surf_short_default, digits = 1, units = "W/m^2^")`
in the default case.

The heat radiated out to space consisted of
`r format_md(I_space_long_asphalt, digits = 3)` W/m^2^ of longwave radiation,
which was 
`r compare_val(I_space_long_asphalt, I_space_long_default, digits = 3, units = "W/m^2^")`
in the default case, and
`r format_md(I_space_short_asphalt, digits = 2)` W/m^2^ of shortwwave radiation,
which was 
`r compare_val(I_space_short_asphalt, I_space_short_default, digits = 2, units = "W/m^2^")`
in the default case.

Then I opened the interactive RRTM model, set the surface type to "Asphalt,"
and manually adjusted the surface temperature until the heat budget was 
balanced.

```{r ex-2-a-balance, include=TRUE, warning=FALSE, message=FALSE}
T_surface_asphalt = 325.6

# Check that the heat really does balance
asphalt_warming = run_rrtm(surface_type = "Asphalt", 
                           T_surface = T_surface_asphalt)
Q_asphalt_warmed = asphalt_warming$Q
Delta_T_asphalt = T_surface_asphalt - T_surface_default
```

The change in I~in~ caused a heat-budget imbalance of 
`r format_md(Q_asphalt, digits = 2)` W/m^2^ going
`r ifelse(Q_asphalt > 0, "into", "out of")` the Earth. 
To put the heat budget back into balance, 
I had to raise the surface temperature 
from `r format_md(T_surface_default, digits = 3)` K to 
`r format_md(T_surface_asphalt, digits = 3)` K, a change of 
`r format_md(Delta_T_asphalt, digits = 2)` K.


#### 2(b) **Concrete Surface**

Next, I did the same for a concrete surface:

```{r ex-2-b, include=TRUE, warning=FALSE, message=FALSE}
concrete = run_rrtm(surface_type = "Concrete")
albedo_concrete = concrete$albedo

I_in_concrete = concrete$i_in
I_out_concrete = concrete$i_out

Q_concrete = concrete$Q

flux_surf_concrete = concrete$fluxes %>% filter(altitude == 0)
flux_space_concrete = tail(concrete$fluxes, 1)

I_surf_long_concrete = flux_surf_concrete$lw_down
I_surf_short_concrete = flux_surf_concrete$sw_down
I_space_long_concrete = flux_space_concrete$lw_up
I_space_short_concrete = flux_space_concrete$sw_up

Delta_I_in_concrete = I_in_concrete - I_in_default
Delta_I_out_concrete = I_out_concrete - I_out_default

Delta_I_surf_long_concrete = I_surf_long_concrete - I_surf_long_default
Delta_I_surf_short_concrete = I_surf_short_concrete - I_surf_short_default
Delta_I_space_long_concrete = I_space_long_concrete - I_space_long_default
Delta_I_space_short_concrete = I_space_short_concrete - I_space_short_default
```

Where asphalt had a much higher albedo albedo for concrete is 
`r format_md(albedo_concrete, digits = 1)` and this
caused I~in~ to decrease to 
`r format_md(I_in_concrete, digits = 3)` W/m^2^.
I~out~ did not change and remained at 
`r format_md(I_out_concrete, digits = 3)` W/m^2^.

Breaking down the heat flow into shortwave and longwave components, the 
heat absorbed by the ground consisted of
`r format_md(I_surf_long_concrete, digits = 3)` W/m^2^ of longwave radiation,
which was 
`r compare_val(I_surf_long_concrete, I_surf_long_default, digits = 1, units = "W/m^2^")`
in the default case, and
`r format_md(I_surf_short_concrete, digits = 3)` W/m^2^ of shortwwave radiation,
which was 
`r compare_val(I_surf_short_concrete, I_surf_short_default, digits = 1, units = "W/m^2^")`
in the default case.

The heat radiated out to space consisted of
`r format_md(I_space_long_concrete, digits = 3)` W/m^2^ of longwave radiation,
which was 
`r compare_val(I_space_long_concrete, I_space_long_default, digits = 2, units = "W/m^2^")`
in the default case, and
`r format_md(I_space_short_concrete, digits = 3)` W/m^2^ of shortwwave radiation,
which was 
`r compare_val(I_space_short_concrete, I_space_short_default, digits = 2, units = "W/m^2^")`
in the default case.


Then I manually adjusted the surface temperature in the interactive RRTM model
until the heat out balanced the heat in.

```{r ex-2-b-balance, include=TRUE, warning=FALSE, message=FALSE}
T_surface_concrete = 247.0

# Check that the heat really does balance
concrete_warming = run_rrtm(surface_type = "Concrete", 
                           T_surface = T_surface_concrete)
Q_concrete_warmed = concrete_warming$Q
Delta_T_concrete = T_surface_concrete - T_surface_default
```

The change in I~in~ caused a heat-budget imbalance of 
`r format_md(Q_concrete, digits = 2)` W/m^2^ going
`r ifelse(Q_concrete > 0, "into", "out of")` the Earth. 
To put the heat budget back into balance, 
I had to lower the surface temperature to 
`r format_md(T_surface_concrete, digits = 3)` K, a change of 
`r format_md(Delta_T_concrete, digits = 2)` K.

#### 2(c) **Discussion**

Covering the earth in asphalt caused the planet to warm up by 
`r format_md(abs(Delta_T_asphalt), digits = 2)` K, 
so we could expect that covering
a large part of a city in asphalt or similarly dark-colored material would
cause the city's temperature to rise.
Conversely, covering the planet in concrete caused it to cool off by 
`r format_md(abs(Delta_T_concrete), digits = 2)` K, so replacing asphalt with 
concrete in a city would cool the city off a good deal.
This would provide cooler temperature for the people living in the city and 
would reduce the impact of global climate change.

## Exercise 3: Clouds

First, run the RRTM model in its default configuration and note the surface
temperature and the albedo.

a) Change the low cloud fraction to 0.70 (70%)
   * Report the changes in shortwave and longwave light absorbed by the 
     surface and going out to space.
   * How much does the total balance of heat change 
     (i.e., how many W/m^2^ does the Earth lose or gain)?
   * Adjust the temperature to bring the heat flows back into balance.
   * How much did the temperature change?

b) Repeat part (a), but with the low cloud fraction set to 0 and the high-cloud
   fraction set to 0.20 (20%).
   
c) Use the `plot_heat_flows()` function to plot the heat flows for the 
   low clouds and the high clouds. Describe the changes you see in the upward 
   and downward heat flows (shortwave, longwave, and total) for the two cases.
   Which kind of cloud had the biggest effect on the outgoing radiation?

### Exercise 3 Report:

#### 3(a) **Low Clouds**

First, I changed the low-cloud fraction to 0.70.

```{r ex-3-a, include=TRUE, warning=FALSE, message=FALSE}
low_clouds = run_rrtm(low_cloud_frac = 0.70)

I_in_low = low_clouds$i_in
I_out_low = low_clouds$i_out

flux_surf_low = low_clouds$fluxes %>% filter(altitude == 0)
flux_space_low = tail(low_clouds$fluxes, 1)

I_surf_long_low = flux_surf_low$lw_down
I_surf_short_low = flux_surf_low$sw_down
I_space_long_low = flux_space_low$lw_up
I_space_short_low = flux_space_low$sw_up

Delta_I_in_low = I_in_low - I_in_default
Delta_I_out_low = I_out_low - I_out_default

Delta_I_surf_long_low = I_surf_long_low - I_surf_long_default
Delta_I_surf_short_low = I_surf_short_low - I_surf_short_default
Delta_I_space_long_low = I_space_long_low - I_space_long_default
Delta_I_space_short_low = I_space_short_low - I_space_short_default

Q_low = low_clouds$Q
```

Adding low clouds caused I~in~ to decrease from 
`r format_md(I_in_default, digits = 3)` W/m^2^ to
`r format_md(I_in_low, digits = 3)` W/m^2^, a change of 
`r format_md(Delta_I_in_low, digits = 2)` W/m^2^.
I~out~ increased from 
`r format_md(I_out_default, digits = 3)` W/m^2^ to
`r format_md(I_out_low, digits = 3)` W/m^2^, a change of 
`r format_md(Delta_I_out_low, digits = 1)` W/m^2^.

Breaking down the heat flow into shortwave and longwave components, the 
heat absorbed by the ground consisted of
`r format_md(I_surf_long_low, digits = 3)` W/m^2^ of longwave radiation,
which was 
`r compare_val(I_surf_long_low, I_surf_long_default, digits = 3, units = "W/m^2^")`
in the default case, and
`r format_md(I_surf_short_low, digits = 3)` W/m^2^ of shortwwave radiation,
which was 
`r compare_val(I_surf_short_low, I_surf_short_default, digits = 3, units = "W/m^2^")`
in the default case.

The heat radiated out to space consisted of
`r format_md(I_space_long_low, digits = 3)` W/m^2^ of longwave radiation,
which was 
`r compare_val(I_space_long_low, I_space_long_default, digits = 3, units = "W/m^2^")`
in the default case, and
`r format_md(I_space_short_low, digits = 3)` W/m^2^ of shortwwave radiation,
which was 
`r compare_val(I_space_short_low, I_space_short_default, digits = 3, units = "W/m^2^")`
in the default case.

Nwxt,  I opened the interactive RRTM model, set the surface type to "low,"
and manually adjusted the surface temperature until the heat budget was 
balanced.

```{r ex-3-a-balance, include=TRUE, warning=FALSE, message=FALSE}
T_surface_low = 280.45

# Check that the new temperature really does balance the heat flow.
low_warming = run_rrtm(low_cloud_frac = 0.70, T_surface = T_surface_low)
Q_low_warmed = low_warming$Q
Delta_T_low = T_surface_low - T_surface_default
```

The change in I~in~ caused a heat-budget imbalance of 
`r format_md(Q_low, digits = 1)` W/m^2^ going
`r ifelse(Q_low > 0, "into", "out of")` the Earth. 
To put the heat budget back into balance, 
I had to lower the surface temperature 
from `r format_md(T_surface_default, digits = 3)` K to 
`r format_md(T_surface_low, digits = 3)` K, a change of 
`r format_md(Delta_T_low, digits = 1)` K.

#### 3(b) **High Clouds**

Next, I changed the high-cloud fraction to 0.70.

```{r ex-3-b, include=TRUE, warning=FALSE, message=FALSE}
high_clouds = run_rrtm(high_cloud_frac = 0.20)
albedo_high = high_clouds$albedo

I_in_high = high_clouds$i_in
I_out_high = high_clouds$i_out

flux_surf_high = high_clouds$fluxes %>% filter(altitude == 0)
flux_space_high = tail(high_clouds$fluxes, 1)

I_surf_long_high = flux_surf_high$lw_down
I_surf_short_high = flux_surf_high$sw_down
I_space_long_high = flux_space_high$lw_up
I_space_short_high = flux_space_high$sw_up

Delta_I_in_high = I_in_high - I_in_default
Delta_I_out_high = I_out_high - I_out_default

Delta_I_surf_long_high = I_surf_long_high - I_surf_long_default
Delta_I_surf_short_high = I_surf_short_high - I_surf_short_default
Delta_I_space_long_high = I_space_long_high - I_space_long_default
Delta_I_space_short_high = I_space_short_high - I_space_short_default

Q_high = high_clouds$Q
```

Adding high clouds caused I~in~ to decrease from 
`r format_md(I_in_default, digits = 3)` W/m^2^ to
`r format_md(I_in_high, digits = 3)` W/m^2^, a change of 
`r format_md(Delta_I_in_high, digits = 2)` W/m^2^.
I~out~ increased from 
`r format_md(I_out_default, digits = 3)` W/m^2^ to
`r format_md(I_out_high, digits = 3)` W/m^2^, a change of 
`r format_md(Delta_I_out_high, digits = 2)` W/m^2^.

Breaking down the heat flow into shortwave and longwave components, the 
heat absorbed by the ground consisted of
`r format_md(I_surf_long_high, digits = 3)` W/m^2^ of longwave radiation,
which was 
`r compare_val(I_surf_long_high, I_surf_long_default, digits = 3, units = "W/m^2^")`
in the default case, and
`r format_md(I_surf_short_high, digits = 3)` W/m^2^ of shortwwave radiation,
which was 
`r compare_val(I_surf_short_high, I_surf_short_default, digits = 3, units = "W/m^2^")`
in the default case.

The heat radiated out to space consisted of
`r format_md(I_space_long_high, digits = 3)` W/m^2^ of longwave radiation,
which was 
`r compare_val(I_space_long_high, I_space_long_default, digits = 3, units = "W/m^2^")`
in the default case, and
`r format_md(I_space_short_high, digits = 3)` W/m^2^ of shortwwave radiation,
which was 
`r compare_val(I_space_short_high, I_space_short_default, digits = 3, units = "W/m^2^")`
in the default case.

Next, I opened the interactive RRTM model, set the surface type to "high,"
and manually adjusted the surface temperature until the heat budget was 
balanced.

```{r ex-3-b-balance, include=TRUE, warning=FALSE, message=FALSE}
T_surface_high = 298.05

# Check that the new temperature really does balance the heat flow.
high_warming = run_rrtm(high_cloud_frac = 0.20, T_surface = T_surface_high)
Q_high_warmed = high_warming$Q
Delta_T_high = T_surface_high - T_surface_default
```

The change in I~in~ caused a heat-budget imbalance of 
`r format_md(Q_high, digits = 2)` W/m^2^ going
`r ifelse(Q_high > 0, "into", "out of")` the Earth. 
To put the heat budget back into balance, 
I had to raise the surface temperature 
from `r format_md(T_surface_default, digits = 3)` K to 
`r format_md(T_surface_high, digits = 3)` K, a change of 
`r format_md(Delta_T_high, digits = 2)` K.

#### 3(c) **Heat Flow**

In the figure below, we can see how the upward and downward heat flows change
with altitude for the case of low clouds.

```{r ex-3-a-figure, fig.cap='Upward and downward fluxes of longwave and shortwave radiation with low clouds.'}
plot_heat_flows(low_clouds)
```

We can see that for the low clouds, the biggest changes are in the downward
flow of radiation. Downward longwave radiation is a lot greater below the clouds
and downward shortwave is smaller. Above the clouds, there is a small increase
in upward longwave and shortwave, but the changes are much smaller than for the
downward radiation.

```{r ex-3-b-figure, fig.cap='Upward and downward fluxes of longwave and shortwave radiation with high clouds.'}
plot_heat_flows(high_clouds)
```

At the height of the clouds we can see big changes in the upward radiation:
Outgoing longwave radiation is considerably smaller above the clouds and 
outgoing shortwave is larger. There are changes to the downward radiation
at the height of the clouds, but these changes are much smaller than for the 
upward radiation.


As these figures show, the low clouds had the greatest effect on
downward radiation and the high clouds had the greatest effect on upward
radiation. In both cases, the clouds affected the longwave radiation much more
than the shortwave.
