---
title: 'Lab #6 Answers'
lab_number: 6
lab_date: '2022-02-28'
sol_date: '2022-03-23'
pubdate: '2022-03-23'
date: '2022-03-16'
pdf_url: /files/lab_solutions/lab_06/lab_06_answers.pdf
slug: lab_06_answers
output:
  blogdown::html_page:
    md_extensions: +tex_math_single_backslash+compact_definition_lists
    toc: yes
---
```{r setup, include=FALSE}
knitr::knit_hooks$set(inline = function(x) { knitr:::format_sci(x, 'md')})
knitr::opts_chunk$set(echo = TRUE, include = TRUE, message = FALSE, 
                      warning = FALSE, cache = FALSE)

if (knitr::is_latex_output()) {
  message("Setting PDF plot hook.")
  new_plot_hook <- function(x, options)
    paste("\n", knitr::hook_plot_tex(x, options), "\n")

  knitr::knit_hooks$set(plot = new_plot_hook)
  knitr::opts_chunk$set(fig.width = 5, fig.height = 5)
} else {
  message("Not using PDF output.")
}


options(dplyr.summarise.inform = FALSE)

# This section loads necessary R libraries and sources scripts that define
# useful functions format_md.
#
data_dir = "_data"
script_dir = "_scripts"

if (!dir.exists(data_dir)) dir.create(data_dir)

library(tidyverse)
library(scales)
library(knitr)

theme_set(theme_bw(base_size = 15))

source(file.path(script_dir, "utils.R"), chdir = T)
source(file.path(script_dir, "format_md.R"), chdir = T)
source(file.path(script_dir, "geocarb.R"), chdir = T)
```
## Exercise 1: Weathering as a Function of CO~2~

In this exercise, I studied how weathering changes when the amount of CO~2~ in
the atmosphere changes, and how this helps the silicate weathering cycle work 
as a thermostat that controls the earth's temperature.

To study changes in weathering, I simulated what would happen if the amount of
volcanic activity around the world suddenly increased, which would mean an 
increase in the amount of volcanic degassing that releases CO~2~ into the 
atmosphere.

The GEOCARB model makes it very easy to simulate a sudden change in degassing:
In class we looked at GEOCARB with an emphasis on what happens after a "spike"
of CO~2~ is suddenly released all at once at year zero. 
For this experiment I was more interested in what happens when there is a 
sustained increase in CO~2~ released to the atmosphere, so I ran GEOCARB with 
the CO~2~ spike set to zero and volcanic degassing that increased from 7.5
during the spinup phase to 10.0 during the simulation phase. 

```{r geocarb-degas}
geocarb_data = run_geocarb(co2_spike = 0, degas_spinup = 7.5, degas_sim = 10.0)
```

With a spike of CO~2~, there was a sudden release at year zero and atmospheric
CO~2~ shot up and then gradually came back down to its former value.  
That happened because there was no sustained source of extra CO~2~. 
But here, when the degassing rate goes up, there is a constant source of extra
CO~2~ so we see something different:

```{r geocarb-degas-plot, fig.cap = "The rise and eventual stabilization of carbon dioxide in the atmophere after a sudden and sustained increase in the rate of volcanic degassing."}
ggplot(geocarb_data, aes(x = year, y = co2_atmos)) + geom_line(size = 1) +
  scale_x_continuous(labels = label_comma()) +
  labs(x = "Year", y = "Atmospheric CO2 (ppm)")
```

As CO~2~ rises, I expected temperature to rise as well, and that is indeed what
I saw.

```{r geocarb-degas-t-plot, fig.cap = "Change in the atmospheric temperature after a sudden and sustained increase in the rate of volcanic degassing of CO2."}
ggplot(geocarb_data, aes(x = year, y = temp_atmos)) + geom_line(size = 1) +
  scale_x_continuous(labels = label_comma()) +
  labs(x = "Year", y = "Atmospheric Temperature (K)")

```

What we see is that when the degassing rate changed, both the 
the CO~2~ concentration and the temperature rise quickly, but as they got
closer to their new equilibrium values, the rate of change slowed down, so that
after about 1 million years, the CO~2~ concentration and temperature were both
slowly approaching their ultimate stable values.

## Stabilization of CO~2~

My first research question was to find out how long it takes CO~2~ to stabilize 
after the degassing rate changes, and what the new stable value is.

The CO~2~ concentration never actually gets to a stable value where it stops 
changing, but over time the rate of change slows down as it gets closer and
closer to the stable value.

This means that we need to define what it means for CO~2~ to stabilize. For the
purposes of this lab, I will follow the lab instructions and define the time
when CO~2~ stabilizes as the time when the rate of change becomes smaller than
some threshold. In this case, I followed the lab instructions and defined
this threshold as a rate of change of 1 part per million in 50,000 years.

I can calculate the rate of change with the R expression
`(co2_atmos - lag(co2_atmos)) / (year - lag(year))`.
I have to include the denominator `(year - lag(year))` because if you look at 
the data returned from the GEOCARB model, the number of years per time step
changes. Right after year 0, the time step between consecutive rows of the 
data is 50 years. At 1,500 years, the time step increases to 500 years per row;
at 30,000 years, it increases to 5,000 years per row;
and at 150,000 years, it increases to 50,000 years per row.

One additional complication is that in this case, the CO~2~ level is rising, but
if I were looking at a case where CO~2~ was falling, the rate of change would
be negative, so even when CO~2~ was changing rapidly 
(e.g., -200 ppm per 50,000 years) the rate of change would
be less than 1 ppm per 50,000 years. 
For this reason, I will define stabilization as occurring when the 
_absolute value_ of the rate of change is less than 1 ppm per 50,000 years.

Taking this working definition of stabilization,I find the year when the 
CO~2~ concentration stabilizes by filtering the model output to keep only the 
rows that meet two conditions:

1. The year is greater than zero (so it's after the degassing rate went up) and 
2. where the absolute value of the rate of change in atmospheric CO~2~ is less 
   than 1 / 50000.

The first row of this new data table will be the year when CO~2~ stabilized,
according to our definition:

```{r geocarb-degas-stabilization}
stabilized = geocarb_data %>% 
  mutate(rate_co2 = (co2_atmos - lag(co2_atmos)) / (year - lag(year))) %>%
  filter(year > 0, abs(rate_co2) < 2E-5)
stabilization_year = stabilized$year[1]
stabilized_co2 = tail(stabilized, 1)$co2_atmos
# Remember that 5E4 is how we tell R 5 times 10^4, which is 50,000
```

After the degassing rate changes, it takes `r stabilization_year` years for the
atmosphere to stabilize. 
Even at 2 million years, CO~2~ is changing slightly, so I estimated the new
stable value of CO~2~ as the value in the last line in the tibble, which 
is `r `round(stabilized_co2, 0)` parts per million, rather than the 
concentration in the year when it stabilized.

## Dynamics of Weathering

Next, I looked at what happened to weathering after the degassing rate 
increased.

```{r geocarb-degas-weathering, fig.cap="Change in the silicate weathering rate after a sudden change in the rate of volcanic degassing of CO2.", fig.width = 6.25}
geocarb_data %>% select(year, degassing_rate, silicate_weathering) %>%
   rename(weathering = "silicate_weathering", degassing = "degassing_rate") %>%
  pivot_longer(cols = -year, names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = year, y = value, color = variable)) +
  geom_line(size = 1) +
  scale_x_continuous(labels = label_comma()) + 
  scale_color_brewer(palette = "Dark2", name = "Rate") +
  labs(x = "Year", y = "Rate (trillion moles/year)")
```

### The Silicate Weathering Thermostat

It can be useful to see how the temperature affects the rate of weathering:

```{r weathering-vs-temp, warning=FALSE}
ggplot(geocarb_data, aes(x = temp_atmos, y = silicate_weathering)) +
   geom_line(size = 1) +
   labs(x = "Atmospheric Temperature (K)", y = "Silicate weathering (trillion moles/year)")
```

There is a clear relationship between temperature and weathering.

It is clear from looking at the figures above that after the degassing rate 
increased:

* The imbalance between degassing and weathering caused the amount of
CO~2~ to rise.
* The rising CO~2~ caused the temeperature to rise.
* And the rising temperature caused weathering to rise.

In the steady state, the rate of weathering must balance the rate of CO~2~
degassing from the Earth, from volcanoes and deep-sea vents, and indeed, as the 
rate of change of CO~2~ in the atmosphere stabilizes, the weathering rate is 
very close to the degassing rate.


The following plot is _**not required for the lab**_, but it can be interesting 
to plot the rate of change in CO~2~ concentration versus the difference between 
degassing and weathering:

```{r co2-vs-weathering, fig.cap = "The rate of change in atmospheric CO2 versus the difference between the degassing rate and the weathering rate", warning = FALSE}
geocarb_data %>% 
   mutate(rate_co2 = (co2_atmos - lag(co2_atmos)) / (year - lag(year)),
          diff = degassing_rate - silicate_weathering) %>%
   ggplot(aes(x = diff, y = rate_co2)) +
   geom_line(size = 1) +
   labs(x = "Degassing - weathering", y = "Rate of change of CO2")
```

In this graph, we can really see the thermostat in action: When there is a big
difference between degassing and weathering, the CO~2~ concentration in the
atmosphere rises, which causes the temperature to rise, which increases the
rate of weathering. As the rate of weathering gets closer to the rate of 
degassing, the difference $(\text{degassing} - \text{weathering}) gets closer
to zero, and the rate of change of CO~2~ becomes smaller. 
Eventually, the atmosphere stabilizes when degassing becomes very close to
weathering, so the difference is very close to zero, and the rate of change of 
CO~2~ becomes so close to zero that for all practical purposes, the amount of
CO~2~ in the atmosphere stops changing.

That is how the silicate weathering thermostat works.

### The effect of changing degassing rates

Finally, I explored how the final, stable CO~2~ concentration and atmospheric
temperature depend on the degassing rate.
Here, I am only interested in the final quasi-stabilized value of CO~2~ and
atmospheric temperature for each value of degassing, so I can see how the 
stable conditions depend on the degassing rate.

```{r vary-degassing, warning=FALSE}
multi_degas_data = tibble()
for (degas_rate in c(7.5, 10, 12.5, 15, 17.5, 20)) {
   geocarb_data = run_geocarb(co2_spike = 0, degas_spinup = 7.5, 
                              degas_sim = degas_rate)
   last_row = tail(geocarb_data, 1)
   multi_degas_data = bind_rows(multi_degas_data, last_row)
}
```

Now I can plot the stable value of CO~2~ versus the degassing rate:

```{r co2_vs_degassing, warning=FALSE}
ggplot(multi_degas_data, aes(x = degassing_rate, y = co2_atmos)) +
   geom_line(size = 1) +
   labs(x = "Degassing (trillion moles per year)", y = "Stabilized CO2 (ppm)")
```

Next, I plot the difference between weathering and degassing at the end of 
each GEOCARB run:

```{r weathering_vs_degassing}
ggplot(multi_degas_data, aes(x = degassing_rate, 
                             y = degassing_rate - silicate_weathering)) +
   geom_line(size = 1) +
   labs(x = "Degassing (trillion moles per year)", y = "Degassing - Weathering")
```

We see here that weathering doesn't exactly match degassing at the end of 
1.95 million years, but it's very close. The differences are less than 1% of
the degassing rate (e.g., a difference of 0.04 when degassing is 16, which 
works out to `r label_percent(accuracy = 0.01)(0.04/16)` of the degassing rate)

Next, I plot the stabilized silicate weathering rate versus CO~2~ concentration

```{r weathering-vs-co2}
ggplot(multi_degas_data, aes(x = co2_atmos, y = silicate_weathering)) +
   geom_line(size = 1) + 
   labs(x = "Stabilized CO2 (ppm)", y = "Silicate weathering (trillion moles/year)")
```

So we see that silicate weathering rises as CO~2~ does. The mechanism for this
is largely the rise in temperature, which we can also plot:

```{r weathering-vs-stable-temp}
ggplot(multi_degas_data, aes(x = temp_atmos, y = silicate_weathering)) +
   geom_line(size = 1) +
   labs(x = "Atmospheric temperature (K)", y = "Silicate weathering (trillion moles/year)")
```

We see that rising temperature strongly affects weathering. This is how the 
silicate weathering thermostat works.


## Exercise 2: The Long-Term Fate of Fossil Fuel CO~2~

In the previous exercise, I looked at how the earth responds to a sustained
increase in CO~2~ emissions over millions of years. Here, I will look at 
what happens when there is a one-time release of a very large quantity of 
CO~2~. From a geological perspective, where people think in terms of millions
of years, this is a good approximation of the release of fossil fuels. In the
end, we expect that people will have burned a lot of fossil fuels in the span
of 200 years or so and then will stop.

Here I am going to investigate what happens after we stop burrning fossil fuels.

I start with a very simple experiment: What would have happened to the 
atmosphere if people didn't burn any fossil fuels. To simulate this, I set
the CO~2~ spike to 0 and also set the volcanic degassing rate during the 
simulation to the same value I used in the spinup (7.5 trillion moles per year).

```{r null-run}
geocarb_null = run_geocarb(co2_spike = 0, degas_spinup = 7.5, degas_sim = 7.5)
```

Now, I plot the change in CO~2~, silicate weathering, carbonate weathering, and
total weathering, when the transition occurs at year 0:

```{r null-run-table}
geocarb_null %>% filter(year >= -200, year <= 200) %>%
   select(year, co2_atmos, silicate_weathering, carbonate_weathering,
          total_weathering) %>%
   kable(digits = 2, 
         col.names = c("Year", "CO~2~", "Silicate Weathering", 
                       "Carbonate Weathering", "Total Weathering"))
```



This is completely unsurprising: If we don't put any extra CO~2~ in the 
atmosphere, the amount of CO~2~ in the atmosphere does not change.

Now I start looking at what happens when we do put CO~2~ in the atmosphere.
I ran GEOCARB with a spike of 2000 billion tons of CO~2~ released into the 
atmosphere at year 0. Near year 0, each time step in the simulation represents 
50 years, so this is a slight simplification, but from the perspective of 
thousands or millions of years, it is pretty close to what we see happening in
the real world.

```{r co2-spike-2000}
geocarb_2k_spike = run_geocarb(co2_spike = 2000, degas_spinup = 7.5, 
                               degas_sim = 7.5)
```

Now I am going to plot what happens to the silicate weathering right after
the spike (i.e., right after year 0), and then I will plot the three kinds of
weathering before and after year zero, when the spike is released.

In the code below, I use the `str_replace_all` function to change
`silicate_weathering` to `silicate` and so forth. I also use the `ordered`
function to arrange the order in which `silicate`, `carbonate`, and `total` 
appear in the legend, from top to bottom, to match the order the three lines 
appear in the graph. That's a level of fussiness I don't expect from students' 
lab reports, but I put it in so you can see how it's done.

```{r plot_co2_spike-fast, fig.cap = "Silicate weathering right after a spike of CO~2~ containing 2000 BT carbon is released"}
weathering_history = geocarb_2k_spike %>% 
   select(year, silicate_weathering, carbonate_weathering, total_weathering) %>%
   pivot_longer(cols = -year, names_to = "Weathering", values_to = "rate") %>%
   mutate(Weathering = str_replace_all(Weathering, "_weathering", "") %>%
             ordered(levels = c("total", "silicate", "carbonate")))

ggplot(weathering_history, aes(x = year, y = rate, color = Weathering)) + 
   geom_line(size = 1, na.rm = TRUE) +
   scale_color_brewer(palette = "Dark2") +
   xlim(-500, 1000) +
   labs(x = "Year", y = "Rate (trillion moles per year)")
```

Next, I look over a much longer time:

```{r plot_co2_spike-slow, fig.cap = "Silicate weathering right after a spike of CO~2~ containing 2000 BT carbon is released"}
ggplot(weathering_history, aes(x = year, y = rate, color = Weathering)) + 
   geom_line(size = 1, na.rm = TRUE) +
   scale_color_brewer(palette = "Dark2") +
   labs(x = "Year", y = "Rate (trillion moles/year)")
```

To test when the CO~2~ stops changing, I use the same criterion I did in the 
previous exercise: when the absolute value of the rate of change is less than 
1 ppm per 50,000 years. The rate of change of CO~2~ here is negative 
(CO~2~ shoots up right after the spike, and then drops over time until it 
returns to close to its original value). 
I define stabilization as when the rate of change is very close to zero, whether
it's positive or negative, so my criterion is for the absolute value of the 
rate of change to be less than 1 ppm per 50,000 years.

The stabilization year is the first year, after year zero, where this is true.
However, because CO~2~ continues to change slowly even after the nominal 
stabilization, I estimate the stable value of CO~2~ to be the value at the very
end of the run, at 1.95 million years.

```{r spike-stabilization}
spike_stabilized = geocarb_2k_spike %>% 
   mutate(co2_rate = (co2_atmos - lag(co2_atmos)) / (year - lag(year))) %>%
   filter(year > 0, abs(co2_rate) < 2E-5)
stabilization_year = head(spike_stabilized$year, 1)
stabilization_value = tail(spike_stabilized$co2_atmos, 1)
```

CO~2~ approximately stabilizes after 
`r label_comma(accuracy = 1000)(stabilization_year)` years, and the final 
stable value is approximately 
`r format_md(stabilization_value, digits = 3)` ppm.

### Weathering and carbonate burial

Next, I looked at how the relationship between total weathering and carbonate 
burial changed over time. Total weathering measures the rate at which new 
carbonate that is released into the oceans, and carbonate burial is the rate
at which carbonate is removed from the oceans by being converted to solid 
minerals and buried on the sea floor.

```{r carbonate-burial-10k}
carbonate_history = geocarb_2k_spike %>% 
   select(year, "Carbon burial" = carbon_burial, 
          "Total weathering" = total_weathering) %>%
   pivot_longer(cols = -year, names_to = "variable", values_to = "value")
ggplot(carbonate_history, aes(x = year, y = value, color = variable)) +
   geom_line(size = 1, na.rm = TRUE) +
   scale_color_brewer(palette = "Dark2", name = NULL) +
   xlim(-1E3, 1E4) +
   labs(x = "Year", y = "Rate (trillion moles per year)")
   

```

When CO~2~ dissolves into the oceans from the atmosphere, it depletes carbonate
in the oceans. This slows down the rate of carbonate burial. Meanwhile, as
CO~2~ builds up in the atmosphere, the temperature rises and the weathering rate
goes up.

Weathering adds more carbonate to the ocean, which replaces the carbonate that
was converted to bicarbonate by reacting with CO~2~ from the atmosphere. 
As the carbonate in the oceans starts to rise, the carbonate burial also rises.

Eventually, as the amount of CO~2~ in the atmosphere stabilizes, the total 
weathering rate and the carbonate burial rate should become roughly equal.

Now, let's look at the rates over a longer period of time:

```{r carbonate-burial-medium}
ggplot(carbonate_history, aes(x = year, y = value, color = variable)) +
   geom_line(size = 1, na.rm = TRUE) +
   scale_color_brewer(palette = "Dark2", name = NULL) +
   xlim(NA, 3E5) +
   labs(x = "Year", y = "Rate (trillion moles per year)")
```

After a few hundred thousand years, the difference between the two rates
is too small to see on a graph that shows the initial spike, so I plotted 
the rates of burial and weathering starting 1 million years after the spike:

```{r carbonate-burial-long}
carbonate_history %>% filter(year >= 1E6) %>%
ggplot(aes(x = year, y = value, color = variable)) +
   geom_line(size = 1, na.rm = TRUE) +
   scale_color_brewer(palette = "Dark2", name = NULL) +
   xlim(1E6, 2E6) +
   labs(x = "Year", y = "Rate (trillion moles per year)")
```


Notice the scale on the vertical axis: The difference between the weathering
rate and the burial rate is less than 0.01 trillion moles per year. This is
a huge difference from 10,000 years after the spike, when the difference was
about 5 trillion moles per year, a factor of 500 greater.

### Summary

In this exercise, I explored different aspects of how the geochemical carbon 
cycle changes after a large amount of CO~2~ is discharged into the air.
What I found was that silicate and carbonate weathering do eventually remove
the excess CO~2~ from the atmosphere and that things stabilize when the 
carbonate burial balances the total weathering.

One important thing this exercise revealed is that the fate of CO~2~
is ultimately controlled by the rate of geological weathering---the rate at
which rain falling on rocks wears them away. This is very slow, and that 
means that even after we stop burning fossil fuels, it can take hundreds of 
thousands of years for CO~2~ to return to its natural concentration in the
atmosphere.


## Exercise 3 (Graduate Students Only): How the Land Plants Changed the Carbon Cycle

In this exercise, I explored how the emergence of land plants altered the 
carbon cycle. To do this, I ran the GEOCARB model with no plants during its
spinup and then added plants for the simulation, so this simulates a condition
where there were no plants for millions of years, and then all of a sudden, at
year zero, lots of land plants appeared.

```{r geocarb-plants}
geocarb_plants = run_geocarb(co2_spike = 0, plants_spinup = FALSE, 
                             plants_sim = TRUE)
```

When the plants appeared, they immediately accelerated the weathering:
```{r weathering-plants}
geocarb_plants %>% filter(year <= 1000) %>% 
   select(year, carbonate_weathering, silicate_weathering, total_weathering) %>%
   pivot_longer(-year, names_to = "weathering", values_to = "rate") %>%
   mutate(weathering = str_replace_all(weathering, "_weathering", "")) %>%
   ggplot(aes(x = year, y = rate, color = weathering)) +
   geom_line(size = 1) +
   scale_color_brewer(palette = "Dark2") +
   scale_x_continuous(labels = label_comma()) +
   labs(x = "Year", y = "Weathering Rate")
```

Over time, we see that the weathering rate gradually drops back to its original 
level.

```{r  weathering-plants-long}
geocarb_plants %>% filter(year <= 1E6) %>%
   select(year, carbonate_weathering, silicate_weathering, total_weathering) %>%
   pivot_longer(-year, names_to = "weathering", values_to = "rate") %>%
   mutate(weathering = str_replace_all(weathering, "_weathering", "")) %>%
   ggplot(aes(x = year, y = rate, color = weathering)) +
   geom_line(size = 1) +
   scale_color_brewer(palette = "Dark2") +
   scale_x_continuous(labels = label_comma()) +
   labs(x = "Year", y = "Weathering Rate")
```

So what does this do to atmospheric CO~2~ concentrations?
When the weathering goes up, our theories tell us that this should remove CO~2~
from the atmosphere by moving it into the ocean as dissolved carbonate ions, 
and ultimately converting it to carbonate rocks, so atmospheric CO~2~ should 
drop.

```{r co2-plants}
geocarb_plants %>% 
   ggplot(aes(x = year, y = co2_atmos)) +
   geom_line(size = 1) +
   scale_x_continuous(labels = label_comma()) +
   labs(x = "Year", y = "Atmospheric CO2 (ppm)")
```

It's useful to show a detail plot of what happens right at the transition where
the plants appear.

```{r co2-plants-short}
geocarb_plants %>% filter(year <= 1000) %>%
   ggplot(aes(x = year, y = co2_atmos)) +
   geom_line(size = 1) +
   scale_x_continuous(labels = label_comma()) +
   labs(x = "Year", y = "Atmospheric CO2 (ppm)")
```

So the CO~2~ starts to drop as soon as the plants appear.
